---
title: "ClimWin"
author: "Nathalie Adenot"
date: "2023-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Import data  

```{r}
library(lubridate)
library(tidyverse)
library(data.table)

# Biological data 
prod <- read.csv("C:/git/STOC/Variables/data/model/Prod.csv")
# Add date column: set by default at 01/07 (format dd/mm/yyyy)
# we set the date at the end of the period we want to consider, at the end of the reproduction period
prod <- prod %>% 
  mutate(ID_PROG = as.factor(ID_PROG), ESPECE = as.factor(ESPECE)) %>% 
  mutate(Date = as.factor(paste("01/07",YEAR, sep = "/")))

## We load data separately because climwin expect daily values, so we don't use the same dataset as in the model
# SPEI 
SPEI <- fread("C:/git/STOC/Variables/data/spei_day1.csv")
SPEI <- SPEI %>% 
  pivot_longer(cols = 2:407, names_to = "ID_PROG", values_to = "SPEI") %>% 
  mutate(YEAR = year(Date)) %>% 
  mutate(Date2 = as.factor(paste(ifelse(day(Date) <10, paste0(0, day(Date)), day(Date)), ifelse(month(Date) < 10, paste0(0, month(Date)), month(Date)), year(Date), sep = "/"))) %>% 
  mutate(ID_PROG = as.factor(ID_PROG)) 
#replace Inf values by value from the previous day
SPEI$SPEI[is.infinite(SPEI$SPEI)] <- NA 
library(zoo)
SPEI <- SPEI %>% 
  arrange(ID_PROG) %>% 
  na.locf()


# NDVI 
NDVI <- fread("C:/git/STOC/Variables/data/ndvi_mod.csv")
NDVI <- NDVI %>% 
  mutate(Date2 = as.factor(paste(ifelse(day(Date) <10, paste0(0, day(Date)), day(Date)), ifelse(month(Date) < 10, paste0(0, month(Date)), month(Date)), year(Date), sep = "/"))) %>% 
  mutate(ID = as.factor(ID)) 
# Environmental variables are expected to be daily values !!!!  TO FINISH  !!!!
NDVI_day <- NDVI %>%
  # select(ID:NDVI, Date2) %>% 
  rename(ID_PROG = ID, YEAR = Year) %>% 
  right_join(SPEI[,c(1,2,4)]) %>% 
  mutate(Year_ID = paste(YEAR, ID_PROG, sep = "_")) %>% 
  group_by(ID_PROG) %>% 
  arrange(Date, by_group = TRUE) %>% 
  filter(yday(Date) > 59 & yday(Date) < 196)
id <- NDVI$Year_ID
NDVI_day <- NDVI_day %>% filter(Year_ID %in% id) # Je comprends pas pourquoi ça marche qu'en le faisant en 2 fois
# complete NDVI values (replace NAs by the next value)
library(zoo)
NDVI_day <- NDVI_day %>% 
  arrange(ID_PROG) %>% 
  na.locf(na.rm = TRUE, fromLast = TRUE) # on finit le 13 juillet au lieu du 15
# write.csv(NDVI_day, "C:/git/STOC/Variables/data/NDVI_day_climwin", row.names = FALSE)


# Temperature  
Temp <- fread("C:/git/STOC/Variables/data/meteo_all_year.csv")
Temp <- Temp %>% 
  filter(mo > 2 & mo < 8) %>%
  select(DATE2, an, T_Q, ID_PROG) %>%
  mutate(Date = as.factor(paste(ifelse(day(DATE2) <10, paste0(0, day(DATE2)), day(DATE2)), ifelse(month(DATE2) < 10, paste0(0, month(DATE2)), month(DATE2)), year(DATE2), sep = "/"))) %>% 
  mutate(ID_PROG = as.factor(ID_PROG))


```

## Test climwin for short distance migrants 

As Climwin takes a long time to run, I will first try to run it using a subset of data: here I test with great tit and blue tit for Pierre-Yves' station.  
I first test with only temperature. 

```{r, warning=FALSE}
library(climwin)
library(lme4)
# select short distance migrants  
prod_short <- prod %>% 
  filter(MIGRATION == "Court") %>% 
  filter(YEAR > 1989)

prod_short_test <- prod_short %>% 
  filter(ID_PROG == 204) %>% 
  filter(ESPECE %in% c("PARMAJ", "PARCAE"))

Temp_test <- Temp %>%
  filter(ID_PROG == 204 & year(DATE2) > 2005) %>% 
  filter(!is.na(T_Q)) %>% 
  distinct() 

str(Temp)

# test PARMAJ 204, temp
MassWin <- slidingwin(xvar = list(Temp = Temp_test$T_Q),  #climate variable = temperature
                      cdate = Temp_test$Date, 
                      bdate = prod_short_test$Date,
                      baseline = glmer(Prod ~ 1 + (1|ESPECE), data = prod_short_test, family = binomial), #null model
                      cinterval = "day", #daily measures (can be "week" or "month")
                      range = c(365, 0), # testing all possible climate windows anywhere up to 150 days before the biological record
                      type = "absolute", refday = c(01, 06), #fixed starting date, on June 1st
                      stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                      func = "lin"
                      # spatial = list(prod_short$ID_PROG, Temp$ID_PROG)) #linear relationship (we can test several e.g. c("lin", "quad"))
)



# print results
head(MassWin[[1]]$Dataset)
# Interpretation: the best climate window is ~60 days before May 20th

MassWin[[1]]$BestModel #coefficients of the best model
# intercept = 4.3498 ; climate = 0.1044

# Plot
MassOutput <- MassWin[[1]]$Dataset
plotdelta(dataset = MassOutput)
plotweights(dataset = MassOutput)
#  boxplots of the start and end point of all climate windows that make up the 95% confidence set
plotwin(dataset = MassOutput)



## SPEI
MassWin <- slidingwin(xvar = list(SPEI = SPEI$SPEI),
                      cdate = SPEI$Date,
                      bdate = prod$Date,
                      # baseline = lm(cbind(AD, JUV+AD) ~ 1, data = prod_short, family = binomial, weights=AD+JUV),
                      baseline = lm(JUV ~ offset(log(AD+0.1)), data = prod_short),
                      cinterval = "day",
                      range = c(150, 0),
                      type = "absolute", refday = c(01, 06),
                      stat = "mean",
                      func = "lin", 
                      spatial = list(prod$ID_PROG, Temp$ID_PROG)) #we can add the list of sites 


```

We can only set the range when we have data: if we consider only spring temperatures, we can not set range = c(365,0).  
But we are mainly interested in spring climate so we will only consider climate variables from 01/03 to 15/07.  

We will run analyses species per species, but I will now try to run climwin using all sites.  

## Combine sites  

As the model takes a very long time to run with all sites, I will first test with only the sites from the Atlantic region:

```{r select atlantic sites}  
reg <- fread("C:/git/STOC_reporting-master/library/reg_biogeo.csv")

reg$ID_PROG <- as.numeric(gsub("\\D", "", reg$NEW.ID_PROG)) #there was only NEW.ID_PROG but we mostly care about the location

sites_Atl <- reg$ID_PROG[which(reg$BIOGEOREF == "Atlantique")]

```

### Temperature

```{r, warning=FALSE}
library(climwin)
library(lme4)
# select species
species <- "SYLATR"
prod_sp <- prod %>% 
  filter(ESPECE == species) %>% 
  filter(YEAR > 1999)# %>% # As it takes a very long time and there were not a lot of stations before 2000 we only use data from this date
  # filter(ID_PROG %in% sites_Atl)

Temp_full <- Temp # save copy of Temp
# Select only Atlantic sites from 2000
Temp <- Temp %>%
  # filter(ID_PROG %in% sites_Atl & year(DATE2) > 1999) %>%
  filter(!is.na(T_Q)) %>%
  distinct()


MassWin <- slidingwin(xvar = list(Temp = Temp$T_Q),  #climate variable = temperature
                      cdate = Temp$Date, 
                      bdate = prod_sp$Date,
                      baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                      cinterval = "day", #daily measures (can be "week" or "month")
                      range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                      type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                      stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                      func = "lin",
                      spatial = list(prod_sp$ID_PROG, Temp$ID_PROG) #linear relationship (we can test several e.g. c("lin", "quad"))
)



# print results
head(MassWin[[1]]$Dataset)
# Interpretation: the best climate window is ~40 days before July 1st

MassWin[[1]]$BestModel #coefficients of the best model

# Plot
MassOutput <- MassWin[[1]]$Dataset
plotdelta(dataset = MassOutput)
plotweights(dataset = MassOutput) # we can be 95% confident that the true ‘best’ model falls within the shaded region.
#  boxplots of the start and end point of all climate windows that make up the 95% confidence set
plotwin(dataset = MassOutput)


```

For PARMAJ, we find a narrow temperature window between 16/05 and 22/05. 
For SYLATR, ~2days around 06/04.  

To test if it is really significant, we do a randomisation test:  

```{r}
# Randomization
MassRand <- randwin(repeats = 5, 
                    xvar = list(Temp = Temp$T_Q),  #climate variable = temperature
                    cdate = Temp$Date, 
                    bdate = prod_sp$Date,
                    baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                    cinterval = "day", #daily measures (can be "week" or "month")
                    range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                    type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                    stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                    func = "lin",
                    spatial = list(prod_sp$ID_PROG, Temp$ID_PROG)) #linear relationship (we can test several e.g. c("lin", "quad"))

# Test for overfitting
pvalue(dataset = MassWin[[1]]$Dataset, datasetrand = MassRand[[1]], 
       metric = "C", #here we have a large dataset with few repeats in randwin()
       sample.size = 23) #Number of years of data

# Plot hist  
MassRand1 <- MassRand[[1]]
MassOutput <- MassWin[[1]]$Dataset
plothist(dataset = MassOutput, datasetrand = MassRand1)  

                   
```
_Plot best model_  

```{r}

# generate BestModel and BestModelData for a single climate window (e.g. the best window from a previous analysis)

MassSingle <- singlewin(xvar = list(Temp = Temp$T_Q),  #climate variable = temperature
                    cdate = Temp$Date, 
                    bdate = prod_sp$Date,
                    baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                    cinterval = "day", #daily measures (can be "week" or "month")
                    range = c(86, 86), # we start from the window we found previously for the best model
                    type = "absolute", 
                    refday = c(01, 07), #fixed starting date, on July 1st
                    stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                    func = "lin",
                    spatial = list(prod_sp$ID_PROG, Temp$ID_PROG)) 

# plot biological response ~ climate variable
plotbest(dataset = MassOutput,
         bestmodel = MassSingle$BestModel,
         bestmodeldata = MassSingle$BestModelData)

# Plot all previous graphs at the same time  
plotall(dataset = MassOutput,
        datasetrand = MassRand,
        bestmodel = MassSingle$BestModel,
        bestmodeldata = MassSingle$BestModelData)

```

### SPEI  

```{r}
library(climwin)
library(lme4)
# select species
species <- "SYLATR"
prod_sp <- prod %>% 
  filter(ESPECE == species) %>% 
  filter(YEAR > 1999) #%>% 
  # filter(ID_PROG %in% sites_Atl) # keep only Atlantic sites

# keep only sites for which we have climate data  
prod_sp <- prod_sp %>% filter(ID_PROG %in% unique(SPEI$ID_PROG))

# filter SPEI data  
# SPEI <- SPEI %>% filter(ID_PROG %in% sites_Atl & YEAR > 1999)


MassWin <- slidingwin(xvar = list(SPEI = SPEI$SPEI),  #climate variable = temperature
                      cdate = SPEI$Date2, 
                      bdate = prod_sp$Date,
                      baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                      cinterval = "day", #daily measures (can be "week" or "month")
                      range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                      type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                      stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                      func = "lin",
                      spatial = list(prod_sp$ID_PROG, SPEI$ID_PROG) #linear relationship (we can test several e.g. c("lin", "quad"))
)

system.time(slidingwin(xvar = list(SPEI = SPEI$SPEI),  #climate variable = temperature
                      cdate = SPEI$Date2, 
                      bdate = prod_sp$Date,
                      baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                      cinterval = "day", #daily measures (can be "week" or "month")
                      range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                      type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                      stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                      func = "lin",
                      spatial = list(prod_sp$ID_PROG, SPEI$ID_PROG) #linear relationship (we can test several e.g. c("lin", "quad"))
))

# Error in cmatrix[i, ] <- cont$xvar[which(cont$cintno$spatial %in% cont$bintno$spatial[i] &  : incorrect number of subscripts on matrix

# print results
head(MassWin[[1]]$Dataset)
# Interpretation: the best climate window is ~60 days before May 20th

MassWin[[1]]$BestModel #coefficients of the best model

# Plot
MassOutput <- MassWin[[1]]$Dataset
plotdelta(dataset = MassOutput)
plotweights(dataset = MassOutput)
#  boxplots of the start and end point of all climate windows that make up the 95% confidence set
plotwin(dataset = MassOutput)


```
_Test significativity_  
```{r}

# Randomization
MassRand <- randwin(repeats = 5, 
                      xvar = list(SPEI = SPEI$SPEI),  #climate variable = temperature
                      cdate = SPEI$Date2, 
                      bdate = prod_sp$Date,
                      baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                      cinterval = "day", #daily measures (can be "week" or "month")
                      range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                      type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                      stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                      func = "lin",
                      spatial = list(prod_sp$ID_PROG, SPEI$ID_PROG))

system.time(randwin(repeats = 5, 
                      xvar = list(SPEI = SPEI$SPEI),  #climate variable = temperature
                      cdate = SPEI$Date2, 
                      bdate = prod_sp$Date,
                      baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                      cinterval = "day", #daily measures (can be "week" or "month")
                      range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                      type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                      stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                      func = "lin",
                      spatial = list(prod_sp$ID_PROG, SPEI$ID_PROG)))

# Test for overfitting
pvalue(dataset = MassWin[[1]]$Dataset, datasetrand = MassRand[[1]], 
       metric = "C", #here we have a large dataset with few repeats in randwin()
       sample.size = 33) #Number of years of data

# Plot hist  
MassRand1 <- MassRand[[1]]
MassOutput <- MassWin[[1]]$Dataset
plothist(dataset = MassOutput, datasetrand = MassRand1)  

```
 
_Plot best model_  

```{r}

# generate BestModel and BestModelData for a single climate window (e.g. the best window from a previous analysis)

MassSingle <- singlewin(xvar = list(SPEI = SPEI$SPEI),  #climate variable = temperature
                    cdate = SPEI$Date2, 
                    bdate = prod_sp$Date,
                    baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                    cinterval = "day", #daily measures (can be "week" or "month")
                    range = c(107, 104), # we start from the window we found previously for the best model
                    type = "absolute", 
                    refday = c(01, 07), #fixed starting date, on July 1st
                    stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                    func = "lin",
                    spatial = list(prod_sp$ID_PROG, SPEI$ID_PROG)) 

# plot biological response ~ climate variable
plotbest(dataset = MassOutput,
         bestmodel = MassSingle$BestModel,
         bestmodeldata = MassSingle$BestModelData)

# Plot all previous graphs at the same time  
plotall(dataset = MassOutput,
        datasetrand = MassRand1,
        bestmodel = MassSingle$BestModel,
        bestmodeldata = MassSingle$BestModelData)

```


### NDVI  
```{r}
library(climwin)
library(lme4)
# select species
species <- "SYLATR"
prod_sp <- prod %>% 
  filter(ESPECE == species) %>% 
  filter(YEAR > 1999) %>% 
  filter(ID_PROG %in% sites_Atl) # keep only Atlantic sites

# filter NDVI data  
NDVI_day <- fread("C:/git/STOC/Variables/data/NDVI_day_climwin")
NDVI_day <- NDVI_day %>% filter(ID_PROG %in% sites_Atl) %>% 
  mutate(YEAR = as.factor(YEAR)) %>% 
  mutate(Date2 = as.factor(paste(ifelse(day(Date) <10, paste0(0, day(Date)), day(Date)), ifelse(month(Date) < 10, paste0(0, month(Date)), month(Date)), year(Date), sep = "/"))) 

# keep only sites for which we have climate data  
prod_sp <- prod_sp %>% filter(ID_PROG %in% unique(NDVI_day$ID_PROG))


MassWin <- slidingwin(xvar = list(NDVI = NDVI_day$NDVI),  #climate variable = temperature
                      cdate = NDVI_day$Date2, 
                      bdate = prod_sp$Date,
                      baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                      cinterval = "day", #daily measures (can be "week" or "month")
                      range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                      type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                      stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                      func = "lin",
                      spatial = list(prod_sp$ID_PROG, NDVI_day$ID_PROG) #linear relationship (we can test several e.g. c("lin", "quad"))
)

# Error in cmatrix[i, ] <- cont$xvar[which(cont$cintno$spatial %in% cont$bintno$spatial[i] &  : incorrect number of subscripts on matrix

# print results
head(MassWin[[1]]$Dataset)
# Interpretation: the best climate window is ~60 days before May 20th

MassWin[[1]]$BestModel #coefficients of the best model

# Plot
MassOutput <- MassWin[[1]]$Dataset
plotdelta(dataset = MassOutput)
plotweights(dataset = MassOutput)
#  boxplots of the start and end point of all climate windows that make up the 95% confidence set
plotwin(dataset = MassOutput)

```


## ACRSCI (rousserole effarvate)  

### Température  

```{r}
species <- "ACRSCI"
prod_sp <- prod %>% 
  filter(ESPECE == species) %>% 
  filter(YEAR > 1999) %>% # As it takes a very long time and there were not a lot of stations before 2000 we only use data from this date
  filter(ID_PROG %in% sites_Atl)

Temp_full <- Temp # save copy of Temp
# Select only Atlantic sites from 2000
Temp <- Temp %>%
  filter(ID_PROG %in% sites_Atl & year(DATE2) > 1999) %>%
  filter(!is.na(T_Q)) %>%
  distinct()


MassWin <- slidingwin(xvar = list(Temp = Temp$T_Q),  #climate variable = temperature
                      cdate = Temp$Date, 
                      bdate = prod_sp$Date,
                      baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                      cinterval = "day", #daily measures (can be "week" or "month")
                      range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                      type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                      stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                      func = "lin",
                      spatial = list(prod_sp$ID_PROG, Temp$ID_PROG) #linear relationship (we can test several e.g. c("lin", "quad"))
)



# print results
head(MassWin[[1]]$Dataset)
# Interpretation: the best climate window is ~40 days before July 1st

MassWin[[1]]$BestModel #coefficients of the best model

# Plot
MassOutput <- MassWin[[1]]$Dataset
plotdelta(dataset = MassOutput)
plotweights(dataset = MassOutput) # we can be 95% confident that the true ‘best’ model falls within the shaded region.
#  boxplots of the start and end point of all climate windows that make up the 95% confidence set
plotwin(dataset = MassOutput)


```

```{r}
# Randomization
MassRand <- randwin(repeats = 5, 
                    xvar = list(Temp = Temp$T_Q),  #climate variable = temperature
                    cdate = Temp$Date, 
                    bdate = prod_sp$Date,
                    baseline = glmer(Prod ~ 1 + (1|YEAR) + (1|ID_PROG), data = prod_sp, family = binomial), #null model
                    cinterval = "day", #daily measures (can be "week" or "month")
                    range = c(120, 0), # testing all possible climate windows anywhere up to 120 days before the biological record
                    type = "absolute", refday = c(01, 07), #fixed starting date, on June 1st
                    stat = "mean", #mean temperatures (we can test several at the same time e.g. c(max, mean))
                    func = "lin",
                    spatial = list(prod_sp$ID_PROG, Temp$ID_PROG)) #linear relationship (we can test several e.g. c("lin", "quad"))

# Test for overfitting
pvalue(dataset = MassWin[[1]]$Dataset, datasetrand = MassRand[[1]], 
       metric = "C", #here we have a large dataset with few repeats in randwin()
       sample.size = 23) #Number of years of data

# Plot hist  
MassRand1 <- MassRand[[1]]
MassOutput <- MassWin[[1]]$Dataset
plothist(dataset = MassOutput, datasetrand = MassRand1)  

```

