---
title: "ECE"
author: "Nathalie Adenot"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Extreme Climatic events  

We want to investigate when climate is really extreme.  
Extreme climatic events "can be defined quantitatively in two ways:
- related to a specific _threshold_ (possibly impact-related);
- related to their _probability of occurrence_" (Ummenhofer et al., 2017)  

But it is not easy to determine impact-related thresholds, so most of the time we "use a frequency distribution of climatic events and designates conditions falling within an arbitrary percentage region (often 5%) at either end of the distribution as extreme" (Wingfield et al., 2017). 

_SPEI_    
For SPEI, specific threshold are defined in the literature : SPEI < -1 is considered outside the range of normal variability, with -2 < SPEI < -1.5 being considered severe drought and SPEI < -2 extreme drought. But this static threshold is dependent on the data as the SPEI is standardized using the distribution of data during a reference period.  

_Temperature_  
For temperature, it is possible to determine threshold from which organisms are more strongly affected (e.g. higher mortality rate) but it is species-dependant (see Danner et al., 2021,  https://doi.org/10.1002/ece3.7194 ; McWhorter et al., 2018, https://doi.org/10.1242/jeb.168930 ; Cunningham et al., 2013). 
But the threshold can also be defined relative to the distribution of data:  
- "number of hot days was defined as the number of days when daily maximum temperature was higher than 30.9°C, the 90th percentile of our daily maximum temperature data in April-August 2005-2010. This value corresponds well with the definition used in human meteorology, i.e. days with >30°C maximum temperature are considered heat days" (Pipoly 2013)  
_NDVI_  
It is less clear whether it would make sense to consider extreme NDVI values, as it is a more stable phenomenon that rely on long-term processes.  

## Examine distribution  
To determine the thresholds we will use, we first need to look at the distribution of our climate variables.  
As we have several sites in different climates, we consider temperature anomalies (else most hot days would probably be in the Mediterranean area).  



```{r distribution}
library(data.table)
library(ggplot2)
library(dplyr)
var <- fread("C:/git/STOC/Variables/data/model/var_climat_daily_anomalies.csv")

# Plot distribution: hist
#############################
##  Temperature anomalies  ##
#############################
gg <- ggplot(var, aes(x=aTemp)) + 
  geom_histogram(binwidth=0.5)
gg

# get 1%, 5% and 10% quantiles 
q1 <- quantile(var$aTemp, probs = 0.01, na.rm = TRUE)
q99 <- quantile(var$aTemp, probs = 0.99, na.rm = TRUE)
q5 <- quantile(var$aTemp, probs = 0.05, na.rm = TRUE)
q95 <- quantile(var$aTemp, probs = 0.95, na.rm = TRUE)
q10  <- quantile(var$aTemp, probs = 0.1, na.rm = TRUE)
q90  <- quantile(var$aTemp, probs = 0.9, na.rm = TRUE)

# Add vertical line to the plot
gg <- gg + geom_vline(aes(xintercept=q1),
            color="red", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q99),
            color="red", linetype="dashed", linewidth=1)
gg <- gg + geom_vline(aes(xintercept=q5),
            color="blue", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q95),
            color="blue", linetype="dashed", linewidth=1)
gg <- gg + geom_vline(aes(xintercept=q10),
            color="green", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q90),
            color="green", linetype="dashed", linewidth=1)
gg 


############
##  SPEI  ##
############
gg <- ggplot(var, aes(x=SPEI)) + 
  geom_histogram(binwidth=0.1)
gg

# get 1%, 5% and 10% quantiles 
q1 <- quantile(var$SPEI, probs = 0.01, na.rm = TRUE)
q99 <- quantile(var$SPEI, probs = 0.99, na.rm = TRUE)
q5 <- quantile(var$SPEI, probs = 0.05, na.rm = TRUE)
q95 <- quantile(var$SPEI, probs = 0.95, na.rm = TRUE)
q10  <- quantile(var$SPEI, probs = 0.1, na.rm = TRUE)
q90  <- quantile(var$SPEI, probs = 0.9, na.rm = TRUE)

# Add vertical line to the plot
gg <- gg + geom_vline(aes(xintercept=q1),
            color="red", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q99),
            color="red", linetype="dashed", linewidth=1)
gg <- gg + geom_vline(aes(xintercept=q5),
            color="blue", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q95),
            color="blue", linetype="dashed", linewidth=1)
gg <- gg + geom_vline(aes(xintercept=q10),
            color="green", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q90),
            color="green", linetype="dashed", linewidth=1)
gg 

####################
## NDVI anomalies ##
####################

aNDVI <- fread("C:/git/STOC/Variables/data/model/var_NDVI_daily_anomalies.csv")
gg <- ggplot(aNDVI, aes(x=daNDVI)) + 
  geom_histogram(binwidth=0.01)
gg

# get 1%, 5% and 10% quantiles 
q1 <- quantile(aNDVI$daNDVI, probs = 0.01, na.rm = TRUE)
q99 <- quantile(aNDVI$daNDVI, probs = 0.99, na.rm = TRUE)
q5 <- quantile(aNDVI$daNDVI, probs = 0.05, na.rm = TRUE)
q95 <- quantile(aNDVI$daNDVI, probs = 0.95, na.rm = TRUE)
q10  <- quantile(aNDVI$daNDVI, probs = 0.1, na.rm = TRUE)
q90  <- quantile(aNDVI$daNDVI, probs = 0.9, na.rm = TRUE)

# Add vertical line to the plot
gg <- gg + geom_vline(aes(xintercept=q1),
            color="red", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q99),
            color="red", linetype="dashed", linewidth=1)
gg <- gg + geom_vline(aes(xintercept=q5),
            color="blue", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q95),
            color="blue", linetype="dashed", linewidth=1)
gg <- gg + geom_vline(aes(xintercept=q10),
            color="green", linetype="dashed", linewidth=1) 
gg <- gg +geom_vline(aes(xintercept=q90),
            color="green", linetype="dashed", linewidth=1)
gg 

```

On these graphs, we don't see a threshold that would really characterize abnormal days.  
10% seems to be still too much within the distribution, so maybe 5% is the best solution.  

For SPEI, the 5% quantile corresponds to -1.43, so close to the -1.5 threshold defining severe drought.   
So 5% could be a good solution for both temperature and SPEI.  

```{r consecutive extreme days}
library(tidyverse)
# Calculer nombre de jours consécutifs au-delà du seuil  
# Pour l'instant je pars sur un seuil de 5%

##########
## SPEI ##
##########
# Pour le SPEI, les jours secs correspondent aux faibles valeurs, donc je garde les valeurs sous le seuil
q5 <- quantile(var$SPEI, probs = 0.05, na.rm = TRUE) # -1.43
# Pour le SPEI je vais plutôt partir sur le seuil de -1.5. Le seuil -2 est souvent trouvé dans la littérature mais en prenant -1.5 (= severe drought) on regroupe ~5% des données, comme la température 
spei <- var[,c("ID_PROG", "Date", "an", "d", "SPEI")]
# keep only non NA values
spei <- spei %>% filter(!is.na(SPEI))
#replace Inf values by value from the previous day
spei$SPEI[is.infinite(spei$SPEI)] <- NA 
library(zoo)
spei <- spei %>% 
  arrange(ID_PROG) %>% 
  na.locf()

# Add column ECE to say if it's an extreme day  
spei$ECE <- FALSE
spei$ECE[spei$SPEI < -1.5] <- TRUE


# Add column with the number of consecutive extreme days
spei$nbECE <- 0

# keep only values from 01/03 to 15/07
spei <- spei[spei$d > 59 & spei$d < 197,]

####  Count nb of ECE per site and year  ####

spei_T <- spei %>% filter(ECE == TRUE) # keep only ECE to have less rows to go through in the loop

spei_nb_ece <- spei_T %>% 
  group_by(ID_PROG, an) %>% 
  count() %>% 
  rename(nb_ECE_SPEI = n)


n <- 1 #initialize number of ECE in a row (we only have ECE so minimum is 1)

#####  Count number of consecutive days  #####
for(i in 2:nrow(spei_T)){
  if(spei_T$ID_PROG[i] != spei_T$ID_PROG[i-1] | spei_T$an[i] != spei_T$an[i-1] | spei_T$d[i] != spei_T$d[i-1] + 1){ #if new site or new year or not consecutive ECE
    n <- 1
    spei_T$nbECE[i] <- n
  } else if(spei_T$d[i] == spei_T$d[i-1] + 1){ # if ECE the next day
     n <- n+1  
     spei_T$nbECE[i] <- n
  } 
}

#get max value  
mx <- spei_T$nbECE[nrow(spei_T)]
for(i in (nrow(spei_T)-1):2){
  if((spei_T$ID_PROG[i] != spei_T$ID_PROG[i+1] | spei_T$an[i] != spei_T$an[i+1])) mx <- spei_T$nbECE[i]
  if(spei_T$nbECE[i] < mx & spei_T$d[i] == (spei_T$d[i+1] - 1)) spei_T$nbECE[i] <- 0
  if(spei_T$nbECE[i] > mx) mx <- spei_T$nbECE[i]
}

spei_ece <- spei %>% 
  group_by(ID_PROG, an, ECE) %>% 
  count() %>% 
  pivot_wider(names_from = ECE, values_from = n) %>% 
  mutate(prop_ECE = TRUE/(TRUE + FALSE))

# write.csv(spei_T, file = "C:/git/STOC/Variables/data/model/nbECEconsSPEI.csv", row.names = FALSE)

# Plot nombre de jours consécutifs au-delà du seuil 
spei_T <- spei_T %>% filter(nbECE>0) %>% mutate(nbECE = as.factor(nbECE))
gg <- ggplot(spei_T, aes(x=nbECE)) + 
  geom_bar() + labs(title = "Number of consecutive days considered ECE (SPEI < -1.5)")
gg


##################
## Températures ##
##################
# Pour les températures, les jours chauds correspondents aux anomalies positives, donc je garde les valeurs au-dessus du seuil
q95 <- quantile(var$aTemp, probs = 0.95, na.rm = TRUE) # 
atemp <- var[,c("ID_PROG", "Date", "an", "d", "aTemp")]
# keep only values from 01/03 to 15/07
atemp <- atemp[atemp$d > 59 & atemp$d < 197,]


# Add column ECE to say if it's an extreme day  
atemp$ECE <- FALSE
atemp$ECE[atemp$aTemp > q95] <- TRUE

atemp_T <- atemp %>% filter(ECE == TRUE) # keep only ECE to have less rows to go through in the loop

####  Count number of ECE per site and year  ####  
atemp_nb_ece <- atemp_T %>% 
  group_by(ID_PROG, an) %>% 
  count() %>% 
  rename(nb_ECE_aTemp = n)

# merge with SPEI  
nb_ece <- atemp_nb_ece %>% 
  left_join(spei_nb_ece)
write.csv(nb_ece, file = "C:/git/STOC/Variables/data/model/nb_ece.csv", row.names = FALSE)

# Add column with the number of consecutive extreme days
atemp_T$nbECE <- 1

n <- 1 #initialize number of ECE in a row (we only have ECE so minimum is 1)

# Count number of consecutive days
for(i in 2:nrow(atemp_T)){
  if(atemp_T$ID_PROG[i] != atemp_T$ID_PROG[i-1] | atemp_T$an[i] != atemp_T$an[i-1] | atemp_T$d[i] != atemp_T$d[i-1] + 1){ #if new site or new year or not consecutive ECE
    n <- 1
    atemp_T$nbECE[i] <- n
  } else if(atemp_T$d[i] == atemp_T$d[i-1] + 1){ # if ECE the next day
     n <- n+1  
     atemp_T$nbECE[i] <- n
  } 
}

# atemp_true <- atemp_T

#get max value  
mx <- atemp_T$nbECE[nrow(atemp_T)]
for(i in (nrow(atemp_T)-1):2){
  if((atemp_T$ID_PROG[i] != atemp_T$ID_PROG[i+1] | atemp_T$an[i] != atemp_T$an[i+1])) mx <- atemp_T$nbECE[i]
  if(atemp_T$nbECE[i] < mx & atemp_T$d[i] == (atemp_T$d[i+1] - 1)) atemp_T$nbECE[i] <- 0
  if(atemp_T$nbECE[i] > mx) mx <- atemp_T$nbECE[i]
}

atemp_T <- atemp_T %>% filter(nbECE>0) %>% mutate(nbECE = as.factor(nbECE))

# write.csv(atemp_T, file = "C:/git/STOC/Variables/data/model/nbECEconsTemp.csv", row.names = FALSE)

# Plot nombre de jours consécutifs au-delà du seuil 
gg <- ggplot(atemp_T, aes(x=nbECE)) + 
  geom_bar() + labs(title = "Number of consecutive days considered ECE", subtitle =  "(Temperature anomalies > 5.1°C ; >95%)")
gg


```


### Early/late periods  

```{r}

##########
## SPEI ##
##########
# Pour le SPEI, les jours secs correspondent aux faibles valeurs, donc je garde les valeurs sous le seuil
q5 <- quantile(var$SPEI, probs = 0.05, na.rm = TRUE) # -1.43
# Pour le SPEI je vais plutôt partir sur le seuil de -1.5. Le seuil -2 est souvent trouvé dans la littérature mais en prenant -1.5 (= severe drought) on regroupe ~5% des données, comme la température 
spei <- var[,c("ID_PROG", "Date", "an", "d", "SPEI")]
# keep only non NA values
spei <- spei %>% filter(!is.na(SPEI))
#replace Inf values by value from the previous day
spei$SPEI[is.infinite(spei$SPEI)] <- NA 
library(zoo)
spei <- spei %>% 
  arrange(ID_PROG) %>% 
  na.locf()

# Add column ECE to say if it's an extreme day  
spei$ECE <- FALSE
spei$ECE[spei$SPEI < -1.5] <- TRUE

# keep only values from 15/03 to 14/05
spei <- spei[spei$d > 73 & spei$d < 197,]


# # Add column with the number of consecutive extreme days
# spei$nbECE <- 0
# 
# 
# ## EARLY  
# # keep only values from 15/03 to 14/05  
# spei_early <- spei[spei$d > 73 & spei$d < 134,]
# 
# ## LATE  
# # keep only values from 15/05 to 15/07  
# spei_late <- spei[spei$d >= 134 & spei$d < 197,]


####  Count nb of ECE per site and year  ####

spei_T <- spei %>% 
  mutate(period = ifelse(d<134, "early", "late")) %>% filter(ECE == TRUE) # keep only ECE to have less rows to go through in the loop

spei_nb_ece <- spei_T %>% 
  group_by(ID_PROG, an, period) %>% 
  count() %>% 
  rename(nb_ECE_SPEI = n) %>% 
  pivot_wider(names_from = period, values_from = nb_ECE_SPEI) %>% 
  mutate(early_nb_ECE_SPEI = replace_na(early, 0), late_nb_ECE_SPEI = replace_na(late, 0)) 




##################
## Températures ##
##################
# Pour les températures, les jours chauds correspondents aux anomalies positives, donc je garde les valeurs au-dessus du seuil
q95 <- quantile(var$aTemp, probs = 0.95, na.rm = TRUE) # 
atemp <- var[,c("ID_PROG", "Date", "an", "d", "aTemp")]
# keep only values from 01/03 to 15/07
atemp <- atemp[atemp$d > 59 & atemp$d < 197,]


# Add column ECE to say if it's an extreme day  
atemp$ECE <- FALSE
atemp$ECE[atemp$aTemp > q95] <- TRUE

atemp_T <- atemp %>% mutate(period = ifelse(d<134, "early", "late")) %>% filter(ECE == TRUE) # keep only ECE to have less rows to go through in the loop

####  Count number of ECE per site and year  ####  
atemp_nb_ece <- atemp_T %>% 
  group_by(ID_PROG, an, period) %>% 
  count() %>% 
  rename(nb_ECE_aTemp = n) %>% 
  pivot_wider(names_from = period, values_from = nb_ECE_aTemp) %>% 
  mutate(early_nb_ECE_aTemp = replace_na(early, 0), late_nb_ECE_aTemp = replace_na(late, 0)) 


# merge with SPEI  
nb_ece <- atemp_nb_ece %>% 
  left_join(spei_nb_ece, by = c("ID_PROG", "an")) %>% 
  select(ID_PROG, an, early_nb_ECE_aTemp, late_nb_ECE_aTemp, early_nb_ECE_SPEI, late_nb_ECE_SPEI)
write.csv(nb_ece, file = "C:/git/STOC/Variables/data/model/nb_ece_early_late.csv", row.names = FALSE)

# Add column with the number of consecutive extreme days
atemp_T$nbECE <- 1

n <- 1 #initialize number of ECE in a row (we only have ECE so minimum is 1)

# Count number of consecutive days
for(i in 2:nrow(atemp_T)){
  if(atemp_T$ID_PROG[i] != atemp_T$ID_PROG[i-1] | atemp_T$an[i] != atemp_T$an[i-1] | atemp_T$d[i] != atemp_T$d[i-1] + 1){ #if new site or new year or not consecutive ECE
    n <- 1
    atemp_T$nbECE[i] <- n
  } else if(atemp_T$d[i] == atemp_T$d[i-1] + 1){ # if ECE the next day
     n <- n+1  
     atemp_T$nbECE[i] <- n
  } 
}

# atemp_true <- atemp_T

#get max value  
mx <- atemp_T$nbECE[nrow(atemp_T)]
for(i in (nrow(atemp_T)-1):2){
  if((atemp_T$ID_PROG[i] != atemp_T$ID_PROG[i+1] | atemp_T$an[i] != atemp_T$an[i+1])) mx <- atemp_T$nbECE[i]
  if(atemp_T$nbECE[i] < mx & atemp_T$d[i] == (atemp_T$d[i+1] - 1)) atemp_T$nbECE[i] <- 0
  if(atemp_T$nbECE[i] > mx) mx <- atemp_T$nbECE[i]
}

atemp_T <- atemp_T %>% filter(nbECE>0) %>% mutate(nbECE = as.factor(nbECE))

# write.csv(atemp_T, file = "C:/git/STOC/Variables/data/model/nbECEconsTemp.csv", row.names = FALSE)

# Plot nombre de jours consécutifs au-delà du seuil 
gg <- ggplot(atemp_T, aes(x=nbECE)) + 
  geom_bar() + labs(title = "Number of consecutive days considered ECE", subtitle =  "(Temperature anomalies > 5.1°C ; >95%)")
gg


```


In the model, we could consider only extreme episodes of more than x days (e.g. 3 consecutive days as for heat waves - canicule).  

So I will now count how many times there has been more than 3 consecutive dry/hot days. 

```{r long dry periods}
############
##  SPEI  ##
############

nbECE_SPEI <- fread("C:/git/STOC/Variables/data/model/nbECEconsSPEI.csv")

# Count number of events >= 3 days  
nbECE3SPEI <- nbECE_SPEI %>% 
  filter(nbECE > 2) %>% 
  group_by(ID_PROG, an) %>% 
  count() %>% 
  rename(ECE3SPEI = n)

# plot
gg <- ggplot(data=nbECE3SPEI, aes(x=an, y=n/length(unique(nbECE3SPEI$ID_PROG)))) +
  geom_bar(stat="identity") +
  labs(title="Mean number of extreme events > 3 days per site for SPEI (spring)", 
         x="", y = "Mean number of extreme dry events > 3 days per site")
gg


###################
##  Temperature  ##
###################

nbECE_temp <- fread("C:/git/STOC/Variables/data/model/nbECEconsTemp.csv")

# Count number of events >= 3 days  
nbECE3temp <- nbECE_temp %>% 
  filter(nbECE > 2) %>% 
  group_by(ID_PROG, an) %>% 
  count() %>% 
  rename(ECE3temp = n)

# plot
gg <- ggplot(data=nbECE3temp, aes(x=an, y=n/length(unique(nbECE3temp$ID_PROG)))) +
  geom_bar(stat="identity") +
  labs(title="Mean number of extreme events > 3 days per site for temperature (spring)", 
         x="", y = "Mean number of extreme hot events > 3 days per site")
gg


# Merge temp and SPEI  
nb3ECE <- nbECE3temp %>% s
  full_join(nbECE3SPEI) %>% 
  mutate(ECE3SPEI = replace_na(ECE3SPEI, 0))

write.csv(nb3ECE, file = "C:/git/STOC/Variables/data/model/nb3ECEcons.csv", row.names = FALSE)
```



We want to know when and where there are the most extreme days:

```{r}
### Explore distribution of ECE

## Temporal 

# SPEI 
spei_ECE <- spei_ECE[spei_ECE$d > 59 & spei_ECE$d < 197,]
# Count ECEs per year and per site 
spei_ECE_ <- spei_ECE %>% 
  group_by(an, ID_PROG) %>% 
  count()
# Count ECEs per site 
spei_ECE_site <- spei_ECE %>% 
  group_by(ID_PROG) %>% 
  count()
# Count ECEs per year
spei_ECE_year <- spei_ECE %>% 
  group_by(an) %>% 
  count()
# Plot number of ECEs per year
gg <- ggplot(data=spei_ECE_year, aes(x=an, y=n)) +
  geom_bar(stat="identity") +
  labs(title="Total number of ECE per year for SPEI (spring)", 
         x="", y = "Total number of days SPEI < -1.5")
gg
# mean per site  
gg <- ggplot(data=spei_ECE_, aes(x=as.factor(an), y=n)) +
  geom_boxplot() +
  labs(title="Number of ECE per year and site for SPEI (spring)", 
         x="", y = "Number of days SPEI < -2")+
  theme(axis.text.x = element_text(angle = 90))
gg


# Temp 
temp_ECE <- temp_ECE[temp_ECE$d > 59 & temp_ECE$d < 197,]
# Count ECEs per year and per site 
temp_ECE_ <- temp_ECE %>% 
  group_by(an, ID_PROG) %>% 
  count()
# Count ECEs per site 
temp_ECE_site <- temp_ECE %>% 
  group_by(ID_PROG) %>% 
  count()
# Count ECEs per year
temp_ECE_year <- temp_ECE %>% 
  group_by(an) %>% 
  count()
# Plot number of ECEs per year
gg <- ggplot(data=temp_ECE_year, aes(x=an, y=n)) +
  geom_bar(stat="identity") +
  labs(title="Total number of ECE per year for Temperature (spring)", 
         x="", y = "Total number of days aTemp > 5")
gg
# mean per site  
gg <- ggplot(data=temp_ECE_, aes(x=as.factor(an), y=n)) +
  geom_boxplot() +
  labs(title="Number of ECE per year and site for Temperature (spring)", 
         x="", y = "Number of days aTemp > 5")+
  theme(axis.text.x = element_text(angle = 90))
gg


## Spatial  
require(rgdal)
require(ggmap)
require(maptools)
library(sf)
require(maps)

# fond de carte
france <- map_data("france")
world1 <- sf::st_as_sf(map('world', plot = FALSE, fill = TRUE))
france <- sf::st_as_sf(map('france', plot = FALSE, fill = TRUE))
mytheme <- theme(plot.title = element_text(face = "bold",size = rel(1.2), hjust = 0.5),
                 panel.background = element_rect(colour = NA),
                 plot.background = element_rect(colour = NA),
                 axis.title = element_text(face = "bold",size = rel(1)),
                 axis.title.y = element_text(angle=90,vjust =2),
                 axis.title.x = element_text(vjust = -0.2),
                 legend.position=NULL)

## SPEI
# Add coordinates
data_model <- read.csv("D:/Stage/STOC/model/data_model.csv")
spei_ECE_site <- spei_ECE_site %>% 
  left_join(data_model[,c("ID_PROG", "LON", "LAT")]) %>% 
  distinct()

# All years
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = spei_ECE_site,aes(LON,LAT, colour=n),size=2)
gg <- gg + scale_color_gradient2(midpoint=210,  low="#fee5d9", mid="#fb6a4a", high="#99000d", space = "Lab")
plot(gg)


## Temperatures
# Add coordinates
data_model <- read.csv("D:/Stage/STOC/model/data_model.csv")
temp_ECE_site <- temp_ECE_site %>% 
  left_join(data_model[,c("ID_PROG", "LON", "LAT")]) %>% 
  distinct()

# All years
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = temp_ECE_site,aes(LON,LAT, colour=n),size=2)
gg <- gg + scale_color_gradient2(midpoint=210,  low="#fee5d9", mid="#fb6a4a", high="#99000d", space = "Lab")
plot(gg)

```




## Year-site anomalies  
I have determined extreme days for each site.  
I also want to know which site-years are extremes. For that, I will use SPEI computed over spring, and average temperature anomalies.  

### Extreme SPEI  

```{r}
SPEI <- fread("C:/git/STOC/Variables/data/spei_month4.csv")
summary(SPEI)
hist(SPEI$SPEI, breaks = 20)

q5 <- quantile(SPEI$SPEI, probs = 0.05, na.rm = TRUE) # -1.59
q8 <- quantile(SPEI$SPEI, probs = 0.08, na.rm = TRUE) # -1.50 --> -1.5 corresponds to 8% of data

# determine ECE  
ECE_spei <- SPEI %>% 
  select(Year, ID_PROG, SPEI) %>% 
  mutate(ECE = ifelse(SPEI < -1.5, "true", "false"))

# % ECE sites per year  
ECE_spei_year <- ECE_spei %>% 
  group_by(Year, ECE) %>% 
  count()
ECE_spei_year_plot <- ECE_spei_year %>% 
  pivot_wider(names_from = ECE, values_from = n) %>% 
  mutate(true = replace_na(true, 0)) %>% 
  mutate(perc_ECE = true/(true + false))

# plot 
gg <- ggplot(data= ECE_spei_year_plot, aes(x=as.factor(Year), y=perc_ECE)) +
  geom_bar(stat="identity") +
  labs(title="% of sites experiencing ECE per year for SPEI (spring)", 
         x="", y = "number of sites which spring SPEI < 1.5") +
  theme(axis.text.x = element_text(angle=90))
gg


# % ECE years per site
ECE_spei_site <- ECE_spei %>% 
  group_by(ID_PROG, ECE) %>% 
  count() 
ECE_spei_site_plot <- ECE_spei_site %>% 
  pivot_wider(names_from = ECE, values_from = n) %>% 
  mutate(true = replace_na(true, 0)) %>% 
  mutate(perc_ECE = true/(true + false))

## Spatial  
require(rgdal)
require(ggmap)
require(maptools)
library(sf)
require(maps)

# fond de carte
france <- map_data("france")
world1 <- sf::st_as_sf(map('world', plot = FALSE, fill = TRUE))
france <- sf::st_as_sf(map('france', plot = FALSE, fill = TRUE))
mytheme <- theme(plot.title = element_text(face = "bold",size = rel(1.2), hjust = 0.5),
                 panel.background = element_rect(colour = NA),
                 plot.background = element_rect(colour = NA),
                 axis.title = element_text(face = "bold",size = rel(1)),
                 axis.title.y = element_text(angle=90,vjust =2),
                 axis.title.x = element_text(vjust = -0.2),
                 legend.position=NULL)

## SPEI
# Add coordinates
data_model <- read.csv("C:/git/STOC/variables/data/model/data_model.csv")
ECE_spei_site_plot <- ECE_spei_site_plot %>% 
  left_join(data_model[,c("ID_PROG", "LON", "LAT")]) %>% 
  distinct()

# All years
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = ECE_spei_site_plot, aes(LON,LAT, colour=perc_ECE),size=2)
gg <- gg + scale_color_gradient2(midpoint=0.1,  low="#fee5d9", mid="#fb6a4a", high="#99000d", space = "Lab")
gg <- gg + labs(colour = "% of years with \n spring SPEI < -1.5")
plot(gg)


```

### Extreme temperature  

```{r}

# Import temperature data  
climate <- fread("C:/git/STOC/Variables/data/model/var_climat_daily_anomalies.csv")

# Calculate mean spring temperature
spring_temp <- climate %>% 
  mutate(d = yday(Date)) %>% 
  filter(d > 73 & d < 197) %>% 
  group_by(an, ID_PROG) %>% 
  summarise(mean_aT = mean(aTemp))
  
summary(spring_temp)
hist(spring_temp$mean_aT, breaks = 20)

q95 <- quantile(spring_temp$mean_aT, probs = 0.95, na.rm = TRUE) # 1.35
q92 <- quantile(spring_temp$mean_aT, probs = 0.92, na.rm = TRUE) # 1.19 --> 8% of data corresponds to +1.19°C

# determine ECE  
ECE_temp <- spring_temp %>% 
  select(an, ID_PROG, mean_aT) %>% 
  mutate(ECE = ifelse(mean_aT > 1.35, "true", "false"))

# % ECE sites per year  
ECE_temp_year <- ECE_temp %>% 
  group_by(an, ECE) %>% 
  count()
ECE_temp_year_plot <- ECE_temp_year %>% 
  pivot_wider(names_from = ECE, values_from = n) %>% 
  mutate(true = replace_na(true, 0)) %>% 
  mutate(perc_ECE = true/(true + false))

# plot 
gg <- ggplot(data= ECE_temp_year_plot, aes(x=as.factor(an), y=perc_ECE)) +
  geom_bar(stat="identity") +
  labs(title="% of sites experiencing ECE per year for temperature (spring)", 
         x="", y = "number of sites which spring temperature anomalies < 1.35") +
  theme(axis.text.x = element_text(angle=90))
gg


# % ECE years per site
ECE_temp_site <- ECE_temp %>% 
  group_by(ID_PROG, ECE) %>% 
  count() 
ECE_temp_site_plot <- ECE_temp_site %>% 
  pivot_wider(names_from = ECE, values_from = n) %>% 
  mutate(true = replace_na(true, 0)) %>% 
  mutate(perc_ECE = true/(true + false))

## Spatial  
require(rgdal)
require(ggmap)
require(maptools)
library(sf)
require(maps)

# fond de carte
france <- map_data("france")
world1 <- sf::st_as_sf(map('world', plot = FALSE, fill = TRUE))
france <- sf::st_as_sf(map('france', plot = FALSE, fill = TRUE))
mytheme <- theme(plot.title = element_text(face = "bold",size = rel(1.2), hjust = 0.5),
                 panel.background = element_rect(colour = NA),
                 plot.background = element_rect(colour = NA),
                 axis.title = element_text(face = "bold",size = rel(1)),
                 axis.title.y = element_text(angle=90,vjust =2),
                 axis.title.x = element_text(vjust = -0.2),
                 legend.position=NULL)

## SPEI
# Add coordinates
data_model <- read.csv("C:/git/STOC/variables/data/model/data_model.csv")
ECE_temp_site_plot <- ECE_temp_site_plot %>% 
  left_join(data_model[,c("ID_PROG", "LON", "LAT")]) %>% 
  distinct()

# All years
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = ECE_temp_site_plot, aes(LON,LAT, colour=perc_ECE),size=2)
gg <- gg + scale_color_gradient2(midpoint=0.07,  low="#fee5d9", mid="#fb6a4a", high="#99000d", space = "Lab")
gg <- gg + labs(colour = "% of years with spring \n temperature anomalies > 1.35")
plot(gg)


```

### Add ECE variables to the df model  

```{r}
df_mod <- fread("C:/git/STOC/Variables/data/model/df_model_final.csv")

# Add binary variables
df_mod <- df_mod %>% 
  left_join(ECE_spei[,c("Year", "ID_PROG", "ECE")], by = c("ID_PROG", "YEAR" = "Year")) %>% 
  rename(ECE_spei = ECE) %>% 
  left_join(ECE_temp[,c("an", "ID_PROG", "ECE")], by = c("ID_PROG", "YEAR" = "an")) %>% 
  rename(ECE_temp = ECE)

# Add continuous variables  
# nb of ECE days  

# nb of consecutive ECE days  
nbECE <- fread("C:/git/STOC/Variables/data/model/nb_ece.csv")
nb3ECEcons <- fread("C:/git/STOC/Variables/data/model/nb3ECEcons.csv") 

df_mod <- df_mod %>% 
  left_join(nbECE, by = c("ID_PROG", "YEAR" = "an")) %>% 
  left_join(nb3ECEcons, by = c("ID_PROG", "YEAR" = "an"))

#replace NAs by 0  
df_mod <- df_mod %>% 
  mutate(nb_ECE_aTemp = replace_na(nb_ECE_aTemp, 0), nb_ECE_SPEI = replace_na(nb_ECE_SPEI, 0)) %>% 
  mutate(ECE3temp = replace_na(ECE3temp, 0), ECE3SPEI = replace_na(ECE3SPEI, 0))  

write.csv(df_mod, file = "C:/git/STOC/Variables/data/model/df_model_final_ECE.csv", row.names = FALSE)
```

