---
title: "THV"
author: "Nathalie Adenot"
date: "2023-05-10"
output: html_document
---


```{r packages} 
library(tidyverse)
library(data.table)
library(ggplot2)
library(DataExplorer)
library(Factoshiny)
library(FactoMineR)
library(BioStatR)
library(ade4)
library(vegan)
library("ggplot2")
library("factoextra")
library(corrplot)
library(RVAideMemoire)
```



```{r import data}

# most captured species 
sp_capt <- fread("C:/git/STOC/Variables/data/sp_capt.csv")

Storchova <- fread("C:/git/STOC/Variables/data/THV/Storchova.csv")

Moussus <- fread("C:/git/STOC/Variables/data/THV/THV_Moussus.csv")  

Gregory <- fread("C:/git/STOC/Variables/data/THV/THV_Gregory2007.csv")
Gregory <- Gregory[,-1]

THV_sp <- fread("C:/git/STOC/Variables/data/THV/THV_all_sp.csv")

TSI <- fread("C:/git/STOC/Variables/data/THV/espece_indicateur_fonctionel_Jiguet2007.csv")
sp_names <- fread("C:/git/STOC/Variables/data/THV/nom_especes.csv")
TSI <- TSI %>% 
  left_join(sp_names, by = c("pk_species" = "SP")) %>% select(nom_sc, sti_europe) %>% 
  filter(!is.na(nom_sc) & !(is.na(sti_europe)))

# Create df with the variables I want to keep  
df_THV <- Storchova %>% 
  # full_join(Moussus) %>% # it is already in THV_sp
  full_join(Gregory) %>% 
  right_join(THV_sp) %>% 
  pivot_longer(cols = Folivore_B:Omnivore_B, names_to = "Diet_Storchova") %>% 
  filter(value == 1) %>%
  mutate(Diet2 = ifelse(Diet_Storchova %in% c("Arthropods_B", "Other invertebrates_B"), "Insectivorous", "Omnivorous")) %>% # this way a species can be designated as both omnivorous and insectivorous --> these species are omnivorous
  select(Species, HABITAT_SP, MIGRATION, Clutch_MEAN, `Broods per year`, Diet2) %>% 
  left_join(TSI, by = c("Species" = "nom_sc")) %>% 
  distinct()

# species that are designated as insectivorous and omnivorous
omni_sp <- df_THV %>% group_by(Species) %>% count() %>% filter(n>1)

df_THV <- df_THV %>% 
  filter(!(Species %in% omni_sp$Species & Diet2 == "Insectivorous"))
```

### Traits of species I keep for the analyses  

```{r}
Storchova33 <- Storchova %>% 
  filter(Species %in% sp_capt$nom_sc[1:33]) %>% 
  select(Species, Folivore_B:Omnivore_B)
```

# Data exploration  

## Check colinearity  

```{r}
plot_correlation(df_THV, cor_args = list("use" = "pairwise.complete.obs")) 

# Quantitative variables  
THV_quant <- df_THV %>% select(Clutch_MEAN, `Broods per year`, TSI)
plot_correlation(THV_quant, cor_args = list("use" = "pairwise.complete.obs"), type="c") 

# Qualitative variables
THV_qual <- df_THV %>% select(HABITAT_SP, MIGRATION, Diet_Storchova, Diet) %>% 
  mutate(Diet2 = ifelse(Diet_Storchova %in% c("Arthropods_B", "Other invertebrates_B"), "Insectivorous", "Omnivorous"))
plot_correlation(THV_qual[,c("HABITAT_SP", "MIGRATION", "Diet2")], cor_args = list("use" = "pairwise.complete.obs"), type="d") 

```
I checked separately quantitative and qualitative variables, but I can also calculate the "rapport de correlation" between quantitative and qualitative variables.  


```{r ex script CEFE}  

#matrice de corr?lation pour les variables num?riques continues
# on utilise spearman parce que la r?partition n'est pas normale
mat <- cor(df_THV[,c(4, 5, 7)],method = "spearman", use ="complete.obs")

# x = var quantitative
# y = var qualitative
df_THV <- as.data.frame(df_THV)

# create correlation matrix
n <- ncol(df_THV)
M<- matrix(0, n, n)
row.names(M) <- colnames(df_THV)
colnames(M) <- colnames(df_THV)

#On ajoute le rapport de corrélation 
for(i in c(2,3,6)){ # quali
  for(j in c(4,5,7)){ # quanti
    r <- BioStatR::eta2(df_THV[,j],df_THV[,i])
    M[i,j] <- r
  }
}

# keep non null values
M <- M[c(2,3,6), c(4,5,7)]


# Correlogramme
corrplot(M, method="color",   
         type="full", 
         addCoef.col = "black",# Ajout du coefficient de corr?lation
         tl.col=c(rep("black", times = 3), rep("red", times = 3)), tl.srt=45, #Rotation des etiquettes de textes
         # Combiner avec le niveau de significativit?
         #p.mat = p.mat, sig.level = 0.01, 
         # Cacher les coefficients de corr?lation sur la diagonale
         diag=TRUE)



```



## PCA  

I want to do a PCA on species traits to see which ones are associated.  
Actually a MCA is more appropriate as most variables are qualitative

_MCA qual var_

```{r}
# Factoshiny  
res <- Factoshiny(df_THV[,c("HABITAT_SP", "MIGRATION", "Diet2")])

# code  
res.MCA<-MCA(df_THV[, c("HABITAT_SP", "MIGRATION", "Diet2")],graph=FALSE)
plot.MCA(res.MCA, choix='var',title="Graphe des variables",col.var=c(1,2,3))
plot.MCA(res.MCA,invisible= 'ind',col.var=c(1,1,2,2,3,3),title="Graphe de l'ACM",label =c('var'))
```


_PCA quantitative variables_  

```{r}
# Factoshiny  
res <- Factoshiny(df_THV[,c("Clutch_MEAN", "Broods per year", "sti_europe")])



### All variables  
## extraction var actives  
df_THV_act <- df_THV[, c("Clutch_MEAN", "Broods per year", "sti_europe")]
df_THV_act <- df_THV_act %>% 
  filter(!is.na(sti_europe))

# PCA

## Colorer par groupe 
df_THV_pca <- df_THV %>% 
  distinct() %>% 
  filter(!is.na(sti_europe)) %>%
  mutate(HABITAT_SP = as.factor(HABITAT_SP), MIGRATION = as.factor(MIGRATION), Diet2 = as.factor(Diet2))


iris.pca <- PCA(df_THV_pca[, - c(1:3,6)], graph = FALSE)

# biplot by group

## Group by migration
png("C:/git/STOC/Variables/output/pca/group_migr.png")
fviz_pca_biplot (iris.pca,
                 col.ind = df_THV_pca$MIGRATION, #palette = "simpsons",
                 addEllipses = TRUE, ellipse.type = "confidence", label = "var",
                 # col.var = factor(c("Temperature", "Temperature", "SPEI", "SPEI", "NDVI", "NDVI", "Temperature", "NDVI", "SPEI", "Temperature", "SPEI", "Temperature", "Temperature", "SPEI", "SPEI")),
                 repel = TRUE,
                 legend.title = list(fill = "Migration"))+
ggpubr::fill_palette("simpsons")#+      # Couleur des individus
dev.off()


# group by diet
png("C:/git/STOC/Variables/output/pca/group_diet.png")
fviz_pca_biplot (iris.pca,
                 col.ind = df_THV_pca$Diet2, #palette = "simpsons",
                 addEllipses = TRUE, ellipse.type = "confidence", label = "var",
                 # col.var = factor(c("Temperature", "Temperature", "SPEI", "SPEI", "NDVI", "NDVI", "Temperature", "NDVI", "SPEI", "Temperature", "SPEI", "Temperature", "Temperature", "SPEI", "SPEI")),
                 repel = TRUE,
                 legend.title = list(fill = "Diet"))+
ggpubr::fill_palette("simpsons")#+      # Couleur des individus
dev.off()



# group by Habitat
png("C:/git/STOC/Variables/output/pca/group_hab.png")
fviz_pca_biplot (iris.pca,
                 col.ind = df_THV_pca$HABITAT_SP, #palette = "simpsons",
                 addEllipses = TRUE, ellipse.type = "confidence", label = "var",
                 # col.var = factor(c("Temperature", "Temperature", "SPEI", "SPEI", "NDVI", "NDVI", "Temperature", "NDVI", "SPEI", "Temperature", "SPEI", "Temperature", "Temperature", "SPEI", "SPEI")),
                 repel = TRUE,
                 legend.title = list(fill = "Habitat"))+
ggpubr::fill_palette("simpsons")#+      # Couleur des individus
dev.off()

 
```

