---
title: "Variables model"
author: "Nathalie Adenot"
date: "2023-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

In this file, I will describe all the process of creating the variables I will use in the models. They are derived from climate data (SPEI, temperature) and NDVI, computed in other scripts.  

```{r load packages}
require(tidyverse)
require(ggplot2)
require(data.table)
```


# Productivity  
To compute productivity, we need the number of Adults (AD) and Juveniles (JUV) of the different species captured each year in each station.  

As in some cases we captured JUV and no AD, we will rather use the % of juveniles captured than the productivity per se, i.e. the number of juveniles produced per adult (JUV/AD). This is also what is done by euRING in the package cesr.  

We will use the columns AD and JUV in a cbind() in binomial glmm, but I will also add a column Prod = nbJUV/nbTOT for visualization. 

```{r productivity}

## import STOC data
dataSTOC <- fread("C:/git/STOC_reporting-master/data_DB/data.csv")

# delete within session recaptures
dataSTOC_recatch <- dataSTOC %>% 
  group_by(DATE, BAGUE) %>% 
  #filter(DATE == min(DATE)) %>% 
  slice(1) %>% # takes the first occurrence if there is a tie
  ungroup()
# 414958 obs

# Count number of JUV and AD captured
STOC <- uniqueSTOC %>% 
#  select(NEW.ID_PROG, ID_PROG, YEAR, AGE_first, ESPECE, HABITAT, MIGRATION) %>% 
  dplyr::filter(AGE_first != "VOL") %>% 
  mutate(AGE_first = as.factor(AGE_first)) %>% 
  group_by(ID_PROG, YEAR, ESPECE, AGE_first, .drop = FALSE) %>% 
  count() %>% 
  pivot_wider(names_from = AGE_first, values_from = n) %>% 
  mutate(ID_PROG = as.factor(ID_PROG), ESPECE = as.factor(ESPECE))

# Add %Juv
STOC <- STOC %>% 
  mutate(Prod = JUV/(JUV+AD))

# Add traits "MIGRATION" and "HABITAT_SP"  
SP_traits <- fread("C:/git/STOC_reporting-master/library/sp.csv") 
  
STOC <- left_join(STOC, SP_traits, by = c("ESPECE" = "SP"))

# write.csv(STOC, file = "C:/git/STOC/Variables/data/model/Prod.csv", row.names = FALSE)

```


# Environmental variables  

I then add the environmental variables (climate variables SPEI and Temperature + vegetation index NDVI).  

We will need both mean value and anomalies to investigate the effect of the interaction. 

## Mean values  

```{r}

## Import NDVI
NDVI <- fread("C:/git/STOC/Variables/data/ndvi_mod.csv")

# Mean spring NDVI
NDVI_mean <- NDVI %>% 
  mutate(JDay = yday(Date)) %>% 
  dplyr::filter(JDay > 90 & JDay <196) %>% 
  group_by(ID, Year) %>% 
  summarise(mNDVI = mean(NDVI)) %>% 
  rename(ID_PROG = ID, YEAR = Year) %>% 
  mutate(ID_PROG = as.factor(ID_PROG))
# There are years without NDVI values for a few sites because they did not meet the quality criteria  

## Import SPEI  
SPEI <- fread("C:/git/STOC/Variables/data/spei_day1.csv")

SPEI_s <- SPEI %>% 
  pivot_longer(cols = 2:407, names_to = "ID_PROG", values_to = "SPEI") %>% 
  mutate(YEAR = year(Date)) %>% 
  mutate(JDay = yday(Date)) %>% 
  arrange(ID_PROG) %>% 
  na.locf() %>% #replace NAs by previous value
  dplyr::filter(JDay > 90 & JDay <196) %>% #01/04 to 15/07
  group_by(ID_PROG, YEAR) %>% 
  summarise(sumSPEI = sum(SPEI), mSPEI = mean(SPEI)) %>%  #sum SPEI values
  mutate(ID_PROG = as.factor(ID_PROG))


## temperature
data_temp <- fread("C:/git/STOC/Variables/data/meteo_all_year.csv")
meanT <- data_temp %>% 
  dplyr::filter(d > 90 & d <196) %>% 
  group_by(an, ID_PROG) %>% 
  summarise(mTemp = mean(T_Q)) %>% 
  rename(YEAR = an) %>% 
  mutate(ID_PROG = as.factor(ID_PROG))

## merge df  
data_mod <- STOC %>% 
  left_join(NDVI_mean, by = c("ID_PROG", "YEAR")) %>% 
  left_join(SPEI_s, by = c("ID_PROG", "YEAR")) %>% 
  left_join(meanT, by = c("ID_PROG", "YEAR")) %>% 
  filter(YEAR > 1989) # we don't have climate data before 1990

## Spatial  
# Add coordinates
coord_sites <- dataSTOC %>% 
  select(ID_PROG, LON, LAT) %>% 
  mutate(ID_PROG = as.factor(ID_PROG)) %>% 
  distinct()

data_mod <- data_mod %>% 
  mutate(ID_PROG = as.factor(ID_PROG)) %>% 
  left_join(coord_sites, by = "ID_PROG")


# write.csv(data_mod, file = "C:/git/STOC/Variables/data/model/data_model.csv", row.names = FALSE)

```


## SPEI mean values: water balance  

The SPEI gives an anomaly as it is standardized for each site. But we also want to know the effect of the mean climate variables and the interactions between mean and anomalies.  

So now we will not consider the SPEI but only the difference between Precipitations and ETP.  

We also think about using the actual evapotranspiration instead of the potential evapotranspiration.  

```{r spatial mean values, echo=FALSE}

meteo <- fread("C:/git/STOC/Variables/data/meteo_all_year.csv")

# Calculate water balance
meteo$BAL <- meteo$PRELIQ_Q - meteo$ETP_Q

# Using actual evapotranspiration (ETa)
meteo$BALa <- meteo$PRELIQ_Q - meteo$EVAP_Q

# mean yearly water balance per site
meteo_mean <- meteo %>% 
  dplyr::filter(d > 90 & d <196) %>% 
  dplyr::rename(YEAR = an) %>%
  group_by(YEAR, ID_PROG) %>% 
  summarise(mBAL = mean(BAL))  

# using ETa
meteo_mean_ETa <- meteo %>% 
  dplyr::filter(d > 90 & d <196) %>% 
  dplyr::rename(YEAR = an) %>% 
  group_by(YEAR, ID_PROG) %>% 
  summarise(mBALa = mean(BALa))

# add to data_mod
data_mod <- fread("C:/git/STOC/Variables/data/model/data_model.csv")
# data_mod <- data_mod %>%
#   left_join(meteo_mean, by = c("YEAR", "ID_PROG")) %>%
#   left_join(meteo_mean_ETa, by = c("YEAR", "ID_PROG"))

# write.csv(data_mod, file = "C:/git/STOC/Variables/data/model/data_model.csv", row.names = FALSE)

# Overall site mean BAL
mean_site <- data_mod %>% 
  group_by(ID_PROG) %>% 
  summarise(meanBAL = mean(mBAL, na.rm = TRUE)) %>% 
  left_join(data_mod[,c("ID_PROG", "LON", "LAT")]) %>% 
  distinct()

# Overall site mean BALa
mean_site_a <- data_mod %>% 
  group_by(ID_PROG) %>% 
  summarise(meanBALa = mean(mBALa, na.rm = TRUE)) %>% 
  left_join(data_mod[,c("ID_PROG", "LON", "LAT")]) %>% 
  distinct()


##########################    Spatial distribution    ########################## 
require(rgdal)
require(ggmap)
require(maptools)
library(sf)
require(maps)
# fond de carte
france <- map_data("france")
world1 <- sf::st_as_sf(map('world', plot = FALSE, fill = TRUE))
france <- sf::st_as_sf(map('france', plot = FALSE, fill = TRUE))
mytheme <- theme(plot.title = element_text(face = "bold",size = rel(1.2), hjust = 0.5),
                 panel.background = element_rect(colour = NA),
                 plot.background = element_rect(colour = NA),
                 axis.title = element_text(face = "bold",size = rel(1)),
                 axis.title.y = element_text(angle=90,vjust =2),
                 axis.title.x = element_text(vjust = -0.2),
                 legend.position=NULL)

## mean BAL 
# All years
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = mean_site,aes(LON,LAT, colour=meanBAL),size=2.5)
gg <- gg + scale_color_gradient2(midpoint=0,  low="#d53e4f", mid="#c1d8e6", high="#023858", space = "Lab")
plot(gg)

## BALa 
# All years
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = mean_site_a,aes(LON,LAT, colour=meanBALa),size=2.5)
gg <- gg + scale_color_gradient2(midpoint=0,  low="#d53e4f", mid="#c1d8e6", high="#023858", space = "Lab")
plot(gg)


#### With raw ETP ####  
# Overall site mean ETP
mean_site_ETP <- meteo %>% 
  dplyr::filter(d > 90 & d <196) %>% 
  group_by(ID_PROG) %>% 
  summarise(meanETP = mean(ETP_Q, na.rm = TRUE), meanETa = mean(EVAP_Q, na.rm = TRUE)) %>% 
  left_join(data_mod[,c("ID_PROG", "LON", "LAT")]) %>% 
  distinct()
# ETP
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = mean_site_ETP,aes(LON,LAT, colour=meanETP),size=2.5)
# gg <- gg + scale_color_gradient2(midpoint=0,  low="#d53e4f", mid="#c1d8e6", high="#023858", space = "Lab")
plot(gg)
# ETa
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = mean_site_ETP,aes(LON,LAT, colour=meanETa),size=2.5)
# gg <- gg + scale_color_gradient2(midpoint=0,  low="#d53e4f", mid="#c1d8e6", high="#023858", space = "Lab")
plot(gg)

# Difference between ETP and ETa  
mean_site_ETP <- mean_site_ETP %>% 
  mutate(diff = meanETP - meanETa)
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = mean_site_ETP,aes(LON,LAT, colour=diff),size=2.5)
gg <- gg + scale_color_gradient(low="#fde67e", high="#f03b20", space = "Lab")
plot(gg)

## Precipitations & Temperature 
mean_site_TP <- meteo %>% 
  group_by(ID_PROG) %>% 
  summarise(sumP = sum(PRELIQ_Q, na.rm = TRUE), meanT = mean(T_Q, na.rm = TRUE)) %>% 
  left_join(data_mod[,c("ID_PROG", "LON", "LAT")]) %>% 
  distinct()
# Plot map
# Precipitation
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = mean_site_TP,aes(LON,LAT, colour=sumP),size=2.5)
gg <- gg + scale_color_gradient2(midpoint=32379,  low="#fc1405", mid="#c1d8e6", high="#023858", space = "Lab")
plot(gg)
# Temperature
# median = 11.288
gg <- ggplot()
gg <- gg + geom_sf(data = world1,fill="white", colour="#7f7f7f", size=0.2)+ geom_sf(data = france,fill="white", colour="#7f7f7f", size=0.5)
gg <- gg + coord_sf(xlim=c(-5,9),ylim=c(41.5,52))
gg <- gg + geom_point(data = mean_site_TP,aes(LON,LAT, colour=meanT),size=2.5)
gg <- gg + scale_color_gradient2(midpoint=11.288,  low="#023858", mid="#c1d8e6", high="#fc1405", space = "Lab")
plot(gg)
```

Here we can see that it rains a lot in Brittany and not much in the Mediterranean region, which makes sens. 


### Check standardization: SPEI vs. scaled water balance anomalies   

We now want to check whether the standardization of the SPEI is concordant with what we would get by standardizing it ourselves (by substracting the local mean).  

As the mean SPEI for each site is not 0, we wanted to check whether it is similar to what we would have when scaling our own variable.  

So I start by adding columns for water balance anomalies, and then scale them.  

```{r check standardization, echo=FALSE}

# ID_PROG is in fact NEW.ID_PROG
meteo <- meteo %>% 
  rename(NEW.ID_PROG = ID_PROG) %>%
  mutate(ID_PROG = as.numeric(gsub("\\D", "", NEW.ID_PROG))) #only keep numeric part


# We start from the daily data: meteo  
# aggregate on julian day (d)

meteo_mBAL <- meteo %>% 
  group_by(ID_PROG, d) %>% 
  summarise_at(c("BAL", "BALa"), mean, na.rm = TRUE) %>% 
  rename(mBAL = BAL) %>% 
  rename(mBALa = BALa)

# Get anomalies
meteo <- meteo %>% 
  left_join(meteo_mBAL, by = c("ID_PROG", "d")) %>%
  mutate(aBAL = BAL - mBAL) %>%
  mutate(aBALa = BALa - mBALa) %>%
  rename(Date = DATE2) %>%
  mutate(ID_PROG = as.factor(ID_PROG)) %>%
  mutate(NEW.ID_PROG = as.factor(NEW.ID_PROG)) %>%
  mutate(Date = as.character(Date))

# Add SPEI
spei <- fread("C:/git/STOC/Variables/data/spei_day1.csv")
spei <- spei %>% 
  pivot_longer(cols = 2:407, names_to = "NEW.ID_PROG", values_to = "SPEI") %>% 
  mutate(ID_PROG = as.numeric(gsub("\\D", "", NEW.ID_PROG))) %>%  
  mutate(Date = as.character(Date)) %>%   
  mutate(ID_PROG = as.factor(ID_PROG)) %>% 
  mutate(NEW.ID_PROG = as.factor(NEW.ID_PROG))
#  left_join(meteo[, c("ID_PROG", "Date", "BAL", "mBALd", "aBAL", "EVAP_Q")]) 

meteo_mod <- meteo %>% 
  select(14,17,4,6,8:11,15,16,18:21) %>% 
  left_join(spei)

# Keep only spring values
spring_meteo_mod <- meteo_mod %>% 
  rename(YEAR = an) %>% 
  mutate(jDay = yday(Date)) %>% 
  filter(jDay > 90 & jDay <196) 

# scale aBAL & aBALa 
s_meteo <- spring_meteo_mod %>% 
  group_by(ID_PROG) %>% 
  mutate(saBAL = scale(aBAL)) %>% 
  mutate(saBALa = scale(aBALa)) %>% 
  filter(!is.infinite(SPEI))

# mean value per year and site
s_meteo_mean <- s_meteo %>% 
  mutate(YEAR = as.factor(YEAR)) %>% 
  mutate(jDay = as.factor(jDay)) %>% 
  group_by(YEAR, ID_PROG) %>% 
  summarise_if(is.numeric, mean, na.rm = TRUE)


# write.csv(s_meteo_mean, file = "C:/git/STOC/Variables/data/model/var_climat_mean.csv", row.names = FALSE)
# write.csv(meteo_mod, file = "C:/git/STOC/Variables/data/model/var_climat_daily.csv", row.names = FALSE)


```

_Correlations between SPEI and scaled water balance anomalies_   

There should be a high correlation between scaled water balance anomalies and SPEI, as SPEI is the standardized water balance for each site.  

```{r correlations SPEI ~ BAL}

# plot
plot(SPEI ~ saBAL, data = s_meteo_mean)
library(DataExplorer)

plot_correlation(s_meteo_mean, cor_args = list("use" = "pairwise.complete.obs"), type="c") 
plot_scatterplot(split_columns(s_meteo_mean)$continuous, by = "SPEI") 

```

SPEI and scaled water balance are higly correlated, so we can use the SPEI without beeing too concerned.  


### Get anomalies  

In the models, we don't want to study the effect of the mean climate variables, but the difference in climate at a given moment compared to the mean site value, and its interaction with the mean site value.  

We don't need to get anomalies for the SPEI (and we already have anomalies for water balance), but we now need anomalies for *temperature* and *NDVI*.  

```{r anomalies, echo=FALSE}
# Import data SPEI & NDVI
s_meteo_mean <- fread("C:/git/STOC/Variables/data/model/var_climat_mean.csv") #all climate variables, averaged /year /site
meteo_mod <- fread("C:/git/STOC/Variables/data/model/var_climat_daily.csv") # all climate variables, /site /day

data_mod <- fread("C:/git/STOC/Variables/data/model/data_model.csv")
# data_mod$mTemp <- NULL

# temperature
data_temp <- fread("C:/git/STOC/Variables/data/meteo_all_year.csv")
spring_temp <- data_temp %>% 
  filter(d > 59 & d <197)  #keep only spring values (01/04 - 14/07)
  
# local means
meanT <- spring_temp %>% 
  # mutate(ID_PROG = as.factor(ID_PROG)) %>% 
  group_by(d, ID_PROG) %>% #mean per day and not per year
  summarise(mTemp = mean(T_Q, na.rm = TRUE))

# merge temperature with the other variables
meteo_mod <- meteo_mod %>% 
  mutate(d = yday(Date)) %>% # Add Julian day column to meteo_mod
    left_join(meanT, by = c("d", "ID_PROG")) 
  # filter(YEAR > 1990)

## Get daily anomalies (for NDVI & Temperatures)
# Temperature anomalies
meteo_mod <- meteo_mod %>% 
  mutate(aTemp = T_Q - mTemp)
# write.csv(meteo_mod, file = "C:/git/STOC/Variables/data/model/var_climat_daily_anomalies.csv", row.names = FALSE)
Temp_mod <- meteo_mod %>% 
  group_by(ID_PROG, an) %>%
  summarise(mdaTemp = mean(aTemp, na.rm =TRUE)) %>%
  rename(YEAR = an) %>%
  filter(YEAR != 2023)
data_mod <- data_mod %>% left_join(Temp_mod)

# NDVI
ndvi <- fread("C:/git/STOC/Variables/data/ndvi_mod.csv")
ndvi$d <- yday(ndvi$Date)
ndvi <- ndvi %>% 
  rename(ID_PROG = ID)
# mean NDVI for each Julian day (period)
meanNDVI <- ndvi %>%  
  group_by(ID_PROG, d) %>% 
  summarise(mNDVI = mean(NDVI, na.rm = TRUE)) %>% 
  filter(d < 200)
# anomalies
meanNDVI <- meanNDVI %>%  
  left_join(ndvi, by = c("ID_PROG", "d")) %>% 
  mutate(daNDVI = NDVI - mNDVI) %>% 
  rename(YEAR = Year) 
# write.csv(meanNDVI, file = "C:/git/STOC/Variables/data/model/var_NDVI_daily_anomalies.csv", row.names = FALSE)

NDVI_mod <- meanNDVI %>% 
  mutate(ID_PROG = as.factor(ID_PROG)) %>% 
  group_by(ID_PROG, YEAR) %>% 
  summarise(mdaNDVI = mean(daNDVI, na.rm = TRUE))

data_mod <- data_mod %>% left_join(NDVI_mod)

## Get annual anomalies
# Temperature  
locmeanTemp <- meteo_mod %>% 
  group_by(ID_PROG) %>% 
  summarise(locTemp = mean(T_Q, na.rm = TRUE))
ymean_meteo <- meteo_mod %>% 
  group_by(ID_PROG, an) %>% 
  summarise(myTemp = mean(T_Q, na.rm = TRUE)) %>% 
  left_join(locmeanTemp) %>% 
  mutate(yaTemp = myTemp - locTemp) %>% 
  rename(YEAR = an)
data_mod <- data_mod %>% 
  left_join(ymean_meteo)

# NDVI 
locmeanNDVI <- ndvi %>% 
  group_by(ID_PROG) %>% 
  summarise(locNDVI = mean(NDVI, na.rm = TRUE))
ymean_NDVI <- ndvi %>% 
  group_by(ID_PROG, Year) %>% 
  summarise(myNDVI = mean(NDVI, na.rm = TRUE)) %>% 
  left_join(locmeanNDVI) %>% 
  mutate(yaNDVI = myNDVI - locNDVI) %>% 
  rename(YEAR = Year)
data_mod <- data_mod %>% 
  left_join(ymean_NDVI)

# select useful columns (I do it afterwards so I'm not sure I have created all the variables at this point of the Rmd)
data_mod_ <- data_mod %>% 
  select(ID_PROG:MIGRATION, LON, LAT, sumSPEI, mSPEI, myTemp, mdaTemp, myNDVI, mdaNDVI)

# write.csv(data_mod_, file = "C:/git/STOC/Variables/data/model/data_model_a2703.csv", row.names = FALSE)

```
I now have 2 types of anomalies:  
- the *anomalies of annual means* (ya... : I first calculate the mean values per year and site, and substract the mean value of a site across all the years)  
- the *mean of anomalies* (mda... : I first calculate anomalies for each day compared to the mean value for this day across years, and then average the anomalies during spring.  

But I was not convinced regarding the mean of anomalies as I was not sure it would give much information.  
So Romain suggested to aggregate anomalies during 2 periods of similar duration (15/03 - 15/05 and 15/05 to 15/07), and to use both mean value and sd in the models.  
There might not be correlation between anomalies in early and late spring, so we should be able to use both variables in the models.  
Mean and sd are probably correlated, but it is easy to decorrelate them using variation coefficient.  

```{r anomalies Romain}

## Anomalies for temperature  
# Calculate mean and sd for temperature for each period: 15/03-15/05 (74-135) & 15/05-15/07 (135-196)
# In meteo_mod we only have temperature from 01/04 to 14/07 so I need to start over again

s_meteo_mean <- fread("C:/git/STOC/Variables/data/model/var_climat_mean.csv") #all climate variables, averaged /year /site
meteo_mod <- fread("C:/git/STOC/Variables/data/model/var_climat_daily.csv") # all climate variables, /site /day

data_mod <- fread("C:/git/STOC/Variables/data/model/data_model_a.csv")
data_mod$mTemp <- NULL

# temperature
data_temp <- fread("C:/git/STOC/Variables/data/meteo_all_year.csv")
spring_temp <- data_temp %>% 
  filter(d > 73 & d <197)  #keep only spring values (15/03 - 15/07)
  
# local means
meanT <- spring_temp %>% 
  # mutate(ID_PROG = as.factor(ID_PROG)) %>% 
  group_by(d, ID_PROG) %>% #mean per day and not per year
  summarise(mTemp = mean(T_Q, na.rm = TRUE))

# merge temperature with the other variables
meteo_mod <- meteo_mod %>% 
  mutate(d = yday(Date)) %>% # Add Julian day column to meteo_mod
    left_join(meanT, by = c("d", "ID_PROG")) 

## Get daily anomalies (for NDVI & Temperatures)
# Temperature anomalies
meteo_mod <- meteo_mod %>% 
  mutate(aTemp = T_Q - mTemp)

# Now it changes: we want 2 distinct periods
meteo_mod <-  meteo_mod %>%  rename(YEAR = an)
# early spring  
Temp_mod_early <- meteo_mod %>% 
  filter(d > 73 & d < 135) %>% #keep 15/03-14/05
  group_by(ID_PROG, YEAR) %>% 
  summarize(maTemp_early = mean(aTemp, na.rm = TRUE)) #mean
Temp_mod_early_sd <- meteo_mod %>% 
  filter(d > 73 & d < 135) %>% #keep 15/03-14/05
  group_by(ID_PROG, YEAR) %>% 
  summarize(sdaTemp_early = sd(aTemp, na.rm = TRUE))#sd
# late spring  
Temp_mod_late <- meteo_mod %>% 
  filter(d >= 135 & d < 197) %>% #keep 15/05-15/07
  group_by(ID_PROG, YEAR) %>% 
  summarize(maTemp_late = mean(aTemp, na.rm = TRUE)) #mean
Temp_mod_late_sd <- meteo_mod %>% 
  filter(d >= 135 & d < 197) %>% 
  group_by(ID_PROG, YEAR) %>% 
  summarize(sdaTemp_late = sd(aTemp, na.rm = TRUE))  #sd

data_mod <- data_mod %>% 
  left_join(Temp_mod_early) %>% 
  left_join(Temp_mod_early_sd) %>% 
  left_join(Temp_mod_late) %>% 
  left_join(Temp_mod_late_sd)



## Anomalies for NDVI 
ndvi <- fread("C:/git/STOC/Variables/data/ndvi_mod.csv")
ndvi$d <- yday(ndvi$Date)
ndvi <- ndvi %>% 
  mutate(ID_PROG = ID)
meanNDVI <- ndvi %>% 
  group_by(ID_PROG, d) %>% 
  summarise(mNDVI = mean(NDVI, na.rm = TRUE))
meanNDVI <- meanNDVI %>% 
  left_join(ndvi, by = c("ID_PROG", "d")) %>% 
  mutate(mdaNDVI = NDVI - mNDVI) %>% 
  rename(YEAR = Year) 

# Now it changes: we want 2 distinct periods
# early spring  
NDVI_early <- meanNDVI %>% 
  filter(d > 73 & d < 135) %>% #keep 15/03-14/05
  group_by(ID_PROG, YEAR) %>% 
  summarize(maNDVI_early = mean(mdaNDVI, na.rm = TRUE)) #mean
NDVI_early_sd <- meanNDVI %>% 
  filter(d > 73 & d < 135) %>% #keep 15/03-14/05
  group_by(ID_PROG, YEAR) %>% 
  summarize(maNDVI_early_sd = sd(mdaNDVI, na.rm = TRUE)) #sd
# late spring  
NDVI_late <- meanNDVI %>% 
  filter(d >= 135 & d < 197) %>% 
  group_by(ID_PROG, YEAR) %>% 
  summarize(maNDVI_late = mean(mdaNDVI, na.rm = TRUE)) #mean
NDVI_late_sd <- meanNDVI %>% 
  filter(d >= 135 & d < 197) %>% 
  group_by(ID_PROG, YEAR) %>% 
  summarize(maNDVI_late_sd = sd(mdaNDVI, na.rm = TRUE))  #sd

data_mod <- data_mod %>% 
  left_join(NDVI_early) %>% 
  left_join(NDVI_early_sd) %>% 
  left_join(NDVI_late) %>% 
  left_join(NDVI_late_sd)

# write.csv(data_mod, file = "C:/git/STOC/Variables/data/model/data_model_all_a.csv", row.names = FALSE)

```

In the end I don't use Romain's anomalies, only the averaged anomalies.  
In the final model, I distinguish early and late anomalies, but I keep the spring variable for mean values (mean NDVI, T and BAL).  

So I will put these different variables in a single dataframe for the models.  

## Final dataframe  

In this dataframe I want:  
- Productivity  
- spring NDVI  
- spring Temperature  
- spring BAL  
- NDVI anomalies early/late  
- Temperature anomalies early/late  
- SPEI early/late

I use the median hatching date 14/05 (Julian day 134) as a limit between early and late breeding period.  


```{r}
# import climate data
climate <- fread("C:/git/STOC/Variables/data/model/var_climat_daily_anomalies.csv") # all climate variables, /site /day
# get mean values  
climate_mean <- climate %>% 
  filter(d > 73 & d < 197) %>% 
  rename("YEAR" = "an") %>% 
  group_by(YEAR, ID_PROG) %>% 
  summarise(mean_T = mean(T_Q, na.rm = TRUE), mean_BAL = mean(BAL, na.rm = TRUE))

# climate 2 periods  
climate_2p <- fread("C:/git/STOC/Variables/data/model/mean_climate_2periods.csv") 
climate_2p <- climate_2p %>% select(ID_PROG, YEAR, mean_aT_early:late_SPEI)

# import NDVI
ndvi <- fread("C:/git/STOC/Variables/data/model/var_NDVI_daily_anomalies.csv")
# calculate mean NDVI per year 
ndvi_mean <- ndvi %>% group_by(YEAR, ID_PROG) %>% 
  summarise(meanNDVI = mean(NDVI, na.rm = TRUE))
# NDVI 2 periods:
ndvi_2p <- fread("C:/git/STOC/Variables/data/model/mean_ndvi_2periods.csv")
ndvi_2p <- ndvi_2p %>% select(YEAR, ID_PROG, mean_aNDVI_early, mean_aNDVI_late)  

# import temperature categories  
temp_cat <- fread("C:/git/STOC/Variables/data/temp_cat_sites.csv")

# import productivity  
data_mod <- fread("C:/git/STOC/Variables/data/model/data_model_a.csv")


# Merge everything  
df_mod <- data_mod %>% 
  select(ID_PROG:MIGRATION) %>% 
  left_join(climate_mean) %>% 
  left_join(temp_cat[, c(1,3)]) %>% 
  left_join(climate_2p) %>% 
  left_join(ndvi_mean) %>% 
  left_join(ndvi_2p)

write.csv(df_mod, file = "C:/git/STOC/Variables/data/model/df_model_final.csv", row.names = FALSE)
```


### Spring variables  
In addition to testing early and late variables, I want to test spring variables to be able to compare with other studies and see if we gain information by using a finer time scale.  

```{r}
# import climate data
climate <- fread("C:/git/STOC/Variables/data/model/var_climat_daily_anomalies.csv") # all climate variables, /site /day

# get mean spring temperature anomalies  
climate_mean <- climate %>% 
  filter(d > 73 & d < 197) %>% 
  rename("YEAR" = "an") %>% 
  group_by(YEAR, ID_PROG) %>% 
  summarise(mean_aT_spring = mean(aTemp, na.rm = TRUE))


# import NDVI
## Anomalies for NDVI 
ndvi <- fread("C:/git/STOC/Variables/data/ndvi_mod.csv")
ndvi$d <- yday(ndvi$Date)
ndvi <- ndvi %>% 
  mutate(ID_PROG = ID)
meanNDVI <- ndvi %>% 
  group_by(ID_PROG, d) %>% 
  summarise(mNDVI = mean(NDVI, na.rm = TRUE))
meanNDVI <- meanNDVI %>% 
  left_join(ndvi, by = c("ID_PROG", "d")) %>% 
  mutate(mdaNDVI = NDVI - mNDVI) %>% 
  rename(YEAR = Year) 

# spring  
NDVI_spring <- meanNDVI %>% 
  filter(d > 73 & d < 197) %>% #keep 15/03-14/05
  group_by(ID_PROG, YEAR) %>% 
  summarize(aNDVI_spring = mean(mdaNDVI, na.rm = TRUE)) #mean

# spring SPEI  
spei_month4 <- fread("C:/git/STOC/Variables/data/spei_month4.csv") # monthly data
spei_month4 <- spei_month4 %>% select(Year, ID_PROG, SPEI) %>% rename(YEAR = Year)


# merge
spring_anomalies <- climate_mean %>% left_join(NDVI_spring) %>% left_join(spei_month4)

write.csv(spring_anomalies, file = "C:/git/STOC/Variables/data/spring_anomalies.csv", row.names = FALSE)

## Add to model df  
var_mod <- fread("C:/git/STOC/Variables/data/model/var_model_final_allsp_scaled.csv")

var_mod <- var_mod %>% left_join(spring_anomalies) %>% 
  mutate(mean_aT_spring = scale(mean_aT_spring), aNDVI_spring = scale(aNDVI_spring), SPEI = scale(SPEI))
write.csv(var_mod, file = "C:/git/STOC/Variables/data/var_model_final_allsp_scaled.csv", row.names = FALSE)

```

### Habitat  

We want to test if the habitat has an influence on the response of birds to drought. 
So I need to add habitat to the df:

```{r}
# import data with habitat
STOC <- fread("C:/git/STOC/Variables/data/data_STOC.csv")

# import model data  (all species)
var_mod <- fread("C:/git/STOC/Variables/data/model/var_model_final_allsp_scaled.csv")

# get the habitat for each site
STOC <- STOC %>% select(ID_PROG, HABITAT) %>% distinct()

# Add to var_model  
var_mod_hab <- var_mod %>% 
  left_join(STOC) 

summary(var_mod_hab)

```
There are 3 habitats: "Aquatique", "STOC-Rozo" and "Terrestre".  
But "STOC-rozo" sites actually correspond to aquatic habitats, but with a specific protocol.  
We can group both as "Aquatique".  

```{r}
# Transform STOC-Rozo in "Aquatique"
var_mod_hab <- var_mod_hab %>% 
  mutate(HABITAT = gsub("STOC-rozo", "Aquatique", HABITAT))%>% 
  mutate(HABITAT = as.factor(HABITAT))

summary(var_mod_hab)

write.csv(var_mod_hab, file = "C:/git/STOC/Variables/data/model/var_model_allsp_scaled_hab.csv", row.names = FALSE)
```

### Phenology  
We also want to test the effect of phenology, to see if the difference in productivity is due to an earlier or a later breeding.  

We want to add phenology as an anomaly as we are not interested in whether the species tends to reproduce early or late, but in whether this particular year reproduction was early or late.  

```{r}
var_mod_hab <- fread("C:/git/STOC/Variables/data/model/var_model_allsp_scaled_hab.csv")

# Import phenology data
pheno <- fread("C:/git/STOC/Variables/data/pheno_conv_model.csv") # 50 species
pheno <- pheno %>% select(mean, YEAR, ESPECE) %>% rename(pheno = mean)
# 1 value per species and per year, Julian day

# Transform phenology in an anomaly 
# get mean fledging date
pheno_anom <- pheno %>% 
  group_by(ESPECE) %>% 
  summarise(mean_pheno = mean(pheno)) 

# get anomalies
pheno <- pheno %>% 
  left_join(pheno_anom, by = "ESPECE") %>% 
  mutate(a_pheno = pheno - mean_pheno) #<0 if earlier, >0 if later than usual

var_mod_pheno <- var_mod_hab %>% 
  right_join(pheno, by = c("YEAR", "ESPECE"))

write.csv(var_mod_pheno, file = "C:/git/STOC/Variables/data/model/var_model_50sp_scaled_pheno.csv", row.names = FALSE)

```




## Life history traits  

After testing the effect of environmental variables, I would like to test the effect of some life history traits:  
- *migration status*  
- *number of clutches*  
- *habitat* (after testing the effect of the site's habitat)  
- *thermal specialisation index*  
We talked about including diet but it might be very correlated with migration status (long-distance migrants are insectivorous)  

I already have migration status and habitat in my dataset, but I need to add the number of clutches and thermal specialisation index.  

```{r}

# import THV Storchova
THV <- fread("C:/git/STOC/Variables/data/THV/Storchova.csv")

# import THV Moussus
THV_Moussus <- fread("C:/git/STOC/Variables/data/THV/THV_Moussus.csv")

nom_especes <- fread("C:/git/STOC/Variables/data/THV/nom_especes.csv", encoding = "Latin-1")


```

For the number of clutches, we have 2 resources: Storchova and Moussus.  
I want to see if they say the same thing :  

```{r}
# Compare number of clutches, Storchova and Moussus  
plot_clutches <- THV_Moussus %>% 
  left_join(THV, by = "Species") %>% 
  select(Species, `Broods per year`, Brood) %>% 
  rename(Moussus = Brood, Storchova = `Broods per year`) %>% 
  pivot_longer(cols = c("Moussus", "Storchova"), names_to = "Source")

gg <- ggplot(data=plot_clutches, aes(x=Species, y=value, fill=Source)) + 
  geom_bar(stat="identity", position=position_dodge()) +
  theme(axis.text.x = element_text(angle=90)) +
  ggtitle("Comparaison nombre de pontes Moussus - Storchova") +
  ylab("Nombre de pontes")
gg
```

It is the same for most species, so I will keep Storchova as there are more species.  


```{r}

# import model data  (all species)
var_mod <- fread("C:/git/STOC/Variables/data/model/var_model_final_allsp_scaled.csv")

# import output metaanalysis
data_meta <- fread("C:/git/STOC/cluster/data/data_metaanalysis.csv")

# check if all species are in Storchova  
setdiff(unique(var_mod$ESPECE), THV$Species)
setdiff(THV$Species, unique(var_mod$ESPECE))
  
# merge THV from different sources with data
THV_all <- var_mod %>% 
  left_join(nom_especes, by = c("ESPECE" = "SP")) %>% 
  left_join(THV, by = c("nom_sc" = "Species")) %>% 
  left_join(THV_Moussus, by = c("nom_sc" = "Species")) %>% 
  rename(Species = nom_sc) %>% 
  dplyr::select(Species, ESPECE, nom_fr, HABITAT_SP, MIGRATION, Clutch_MEAN, `Broods per year`, TSI) %>% 
  filter(!is.na(MIGRATION)) %>% 
  distinct()

write.csv(THV_all, file = "C:/git/STOC/Variables/data/THV/THV_all_sp.csv", row.names = FALSE)

# merge with meta_data
data_meta_THV <- data_meta %>% 
  mutate(phylo = gsub(pattern = "Parus_caeruleus", "Cyanistes_caeruleus", phylo)) %>% 
  mutate(Species = gsub("_", " ", phylo)) %>% 
  left_join(THV_all, by = "Species") 

write.csv(data_meta_THV, file = "C:/git/STOC/Variables/data/THV/data_meta_THV_all_sp.csv", row.names = FALSE)

```

### TSI  

Source: Jiguet 2007  

```{r}
sti <- fread("C:/git/STOC/Variables/data/THV/espece_indicateur_fonctionel_Jiguet2007.csv")

THV <- fread("C:/git/STOC/Variables/data/THV/THV_all_sp.csv")  

THV <- THV %>% 
  left_join(sti[,c("pk_species", "sti_europe")], by = c("ESPECE" = "pk_species"))

write.csv(THV, file = "C:/git/STOC/Variables/data/THV/THV_all_sp.csv", row.names = FALSE)

```

