---
title: "2bioperiods_models"
author: "Nathalie Adenot"
date: "29/03/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r libraries}
library(data.table)
library(tidyverse)
library(lubridate)
library(DataExplorer)
library(lme4)
library(nlme)
library(mgcv) #gamm
```

## Determine bioperiods  

### Get Life History Traits  
We decided to set the beginning of the *late breeding period* at hatching.  To find this date, we will start from the peak of captured fledglings and substract the number of days it takes for a juvenile to go from hatching to being independant.  
We can find this data in Storchova : 

```{r nestling period}
# import Storchova LHT 
Storchova <- fread("C:/git/STOC/Variables/data/THV/Storchova.csv")

# Import STOC species  
sp_capt <- fread("C:/git/STOC/Variables/data/sp_capt.csv")

# Merge data  
sp_capt_LHT <- sp_capt %>% 
  rename(Species = nom_sc) %>% 
  left_join(Storchova[, c("Species", "Nestling period", "Parental feeding", "Age of independence")])

# write.csv(sp_capt_LHT, file = "C:/git/STOC/Variables/data/THV/STOC_Storchova.csv", row.names = FALSE)


```

There are still some missing values that we can find elsewhere, such as in the *Handbook of the birds of the world*.  
(For other LHT there is AVONET, Tobias 2022 but there is no data on breeding). 

For the rest of the analyses, we will only keep species that met the convergence criteria to compute the phenology indicator:

```{r}

# Import pheno data
pheno <- fread("C:/git/STOC/Variables/data/pheno33.csv")

# Keep only species for which we have the phenology
sp_mod_LHT <- sp_capt_LHT %>% filter(ESPECE %in% pheno$ESPECE)

spNA <- sp_mod_LHT$Species[is.na(sp_mod_LHT$`Nestling period`)]
```

We still need to find data for the species `r spNA`.   

_Add missing data_  

For the species we didn't find in Storchova, I will get the information in the *Handbooks of birds of the world* : 
- *Luscinia svecica* (Bluethroat, Volume 10): "nestling period 13-14 days ; post-fledging dependance up to 1 month, probably usually shorter"
- *Saxicola torquata* (Common stonechat, Volume 10): "nestling period 13-16 days ; post-fledging dependance 3-4 months in Africa, 8-14 days in Europe in cases of first broods" 
- *Carduelis cannabina* (Common linnet, Volume 15): "nestling period 10-14 days, young independant after a further 14 days"  
- *Carduelis chloris* (European greenfinch, Volume 15): "nestling period 14-18 days, young of first brood fed up to 14 days after fledging"   


I will now add this to the dataframe sp_mod_LHT. As we have only an interval, I will keep the median of this interval (and in case of a non integer median, I will round the value):  

```{r}

sp_mod_LHT$`Nestling period`[which(sp_mod_LHT$Species == "Luscinia svecica")] <- round(median(13:14))  
sp_mod_LHT$`Nestling period`[which(sp_mod_LHT$Species == "Saxicola torquata")] <- round(median(13:16))  
sp_mod_LHT$`Nestling period`[which(sp_mod_LHT$Species == "Carduelis cannabina")] <- round(median(10:14))  
sp_mod_LHT$`Nestling period`[which(sp_mod_LHT$Species == "Carduelis chloris")] <- round(median(14:18))  

```

### Get hatching date  

We use the phenology indicator created by Paul to get the mean fledging date of species (1 value /species /year).  
To do so, we will do: *Hatching date = fledgling date - 7 - nestling period*  
The 7 corresponds to the time juveniles need to really start flying around and get captured. We could also use the variables "Age of independance" or "Parents feeding" but juveniles already get captured when they are still flying with their parents.  


```{r pheno}

# Import pheno data
pheno <- fread("C:/git/STOC/Variables/data/pheno.csv")
summary(pheno)
pheno <- pheno %>% 
  mutate(YEAR = as.factor(YEAR), ESPECE = as.factor(ESPECE))

# Add nestling period
pheno <- pheno %>% 
  left_join(sp_mod_LHT) %>% 
  # Calculate hatching date
  mutate(Hatching = round(mean_c - 7 - `Nestling period`))

# write.csv(pheno, file = "C:/git/STOC/Variables/data/pheno_hatching.csv", row.names = FALSE)

```

### Estimate climate variables for each period  

The difficulty here is that climate variable will vary for each species as the periods will differ.  

```{r}
# Import pheno dates  
pheno_dates <- fread("C:/git/STOC/Variables/data/pheno_hatching.csv")

# Import climate data  
climate <- fread("C:/git/STOC/Variables/data/model/var_climat_daily_anomalies.csv")
# Import NDVI
ndvi <- fread("C:/git/STOC/Variables/data/model/var_NDVI_daily_anomalies.csv")

# Merge climate with dates  
climate_pheno <- climate %>% 
  rename(YEAR = an) %>% 
  filter(d > 59 & d <197) %>% # keep spring data
  dplyr::select(ID_PROG, Date, d, YEAR, T_Q, aTemp, BAL, SPEI) 
#replace Inf values by value from the previous day
climate_pheno$SPEI[is.infinite(climate_pheno$SPEI)] <- NA 
# In the end I don't do this because there are sites with many missing values in a row so it messed up everything
# library(zoo)
# climate_pheno <- climate_pheno %>% 
#   group_by(ID_PROG) %>% 
#   na.locf() 
climate_pheno_ <- climate_pheno %>%
  left_join(pheno_dates, by = "YEAR")

# Create variable "early" and "late"  
climate_pheno_p <- climate_pheno_ %>% 
  mutate(period = ifelse(d < Hatching, "early", "late")) %>% 
  group_by(ID_PROG, YEAR, ESPECE, period) %>% 
  summarize(meanBAL = mean(BAL), meanSPEI = mean(SPEI, na.rm =TRUE), meanT = mean(T_Q), mean_aT = mean(aTemp))
  
# wider  
climate_pheno_wide <- climate_pheno_p %>% 
  filter(!is.na(period)) %>% 
  pivot_wider(names_from = period, values_from = c(meanBAL, meanSPEI, meanT, mean_aT))

# write.csv(climate_pheno_wide, file = "C:/git/STOC/Variables/data/model/mean_climate_sp_2periods.csv", row.names = FALSE)

# Merge NDVI with dates  
ndvi_pheno <- ndvi %>% 
  filter(d > 59 & d <197) %>% # keep spring data (normalement on a déjà que le printemps mais bon)
  dplyr::select(ID_PROG, Date, d, YEAR, NDVI, daNDVI) %>% 
# Add pheno
  left_join(pheno_dates, by = "YEAR")  %>% 
# Create variables "early" and "late"  
  mutate(period = ifelse(d < Hatching, "early", "late")) %>% 
  group_by(ID_PROG, YEAR, ESPECE, period) %>% 
  summarize(meanNDVI = mean(NDVI, na.rm = TRUE), mean_aNDVI = mean(daNDVI, na.rm = TRUE)) %>%  
  filter(!is.na(period)) %>% 
  pivot_wider(names_from = period, values_from = c(meanNDVI, mean_aNDVI))

# write.csv(ndvi_pheno, file = "C:/git/STOC/Variables/data/model/mean_ndvi_sp_2periods.csv", row.names = FALSE)

# merge df  
mean_2periods <- merge(climate_pheno_wide, ndvi_pheno, all = TRUE)

# write.csv(mean_2periods, file = "C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv", row.names = FALSE)

```




### Check colinearity  
We want to see if there is colinearity between the variables for early and late spring.  

```{r check colinearity}  
library(DataExplorer)

mean_2periods <- fread("C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv")
ndvi_2periods <- fread("C:/git/STOC/Variables/data/model/mean_ndvi_sp_2periods.csv")

plot_correlation(mean_2periods[,4:13], cor_args = list("use" = "pairwise.complete.obs"), type="c") 

# SPEI
gg <- ggplot(mean_2periods, aes(y = meanSPEI_late, x = meanSPEI_early)) +
  geom_point(stat = "identity")
plot(gg) 
# BAL
gg <- ggplot(mean_2periods, aes(y = meanBAL_late, x = meanBAL_early)) +
  geom_point(stat = "identity")
plot(gg) 

# Temp
gg <- ggplot(mean_2periods, aes(y = meanT_late, x = meanT_early)) +
  geom_point(stat = "identity")
plot(gg) 
# Temp anomalies  
gg <- ggplot(mean_2periods, aes(y = mean_aT_late, x = mean_aT_early)) +
  geom_point(stat = "identity")
plot(gg)

# NDVI  
gg <- ggplot(mean_2periods, aes(y = meanNDVI_late, x = meanNDVI_early)) +
  geom_point(stat = "identity")
plot(gg)
# NDVI anomalies  
gg <- ggplot(mean_2periods, aes(y = mean_aNDVI_late, x = mean_aNDVI_early)) +
  geom_point(stat = "identity")
plot(gg)

```
There is a clear relationship for Temperature, and a pattern for NDVI.  But there is no correlation for SPEI and anomalies.  
So for anomalies I can use both in the same model. 
For the mean temperature and NDVI, the correlation is rather high so we can consider that including only the global spring temperature / NDVI is enough and captures the variation. Moreover, we only include the mean site value to take into account the inter-site variability. So to answer our question using the 2 periods only for anomalies would make sense.  




## Models  

As the variables in early and late breeding season are highly correlated, we will need to detrend them by using the residuals. But I will first explore the relationships between productivity and these variables using univariate models. I use GAMM to explore linearities.  

### GAMM univariate   

#### SPEI + BAL  
_Early spring_  

```{r}

prod <- fread("C:/git/STOC/Variables/data/model/data_model.csv")
var <- fread("C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv")
var <- mean_2periods
var_mod <- prod %>% left_join(var) %>% 
  filter(!is.na(meanSPEI_early)) # now we keep only our 33 species

# write.csv(var_mod, file = "C:/git/STOC/Variables/data/model/data_model_periods.csv")

var_mod <- fread( "C:/git/STOC/Variables/data/model/data_model_periods.csv")

# plot SPEI
sp <- ggplot(data=var_mod, aes(x=meanSPEI_early, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

# select only years >2000 and species in pheno (20 sp)
var_mod2000 <- var_mod %>% filter(YEAR > 1999 & ESPECE %in% pheno$ESPECE)
# Model
mgamm_early_spei <- gamm(cbind(JUV, AD) ~ s(meanSPEI_early), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), 
                        family = binomial, data = var_mod2000)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_spei, file="C:/git/STOC/stats/models/mgamm_early_spei.rda", compress='xz')

plot(mgamm_early_spei$gam,pages=1)
summary(mgamm_early_spei$lme) # details of underlying lme fit
summary(mgamm_early_spei$gam) # gam style summary of fitted model
anova(mgamm_early_spei$gam) 
gam.check(mgamm_early_spei$gam) # simple checking plots


mgamm_early_spei_noSp <- gamm(cbind(JUV, AD) ~ s(meanSPEI_early), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_spei_noSp, file="C:/git/STOC/stats/models/mgamm_early_spei_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_spei_noSp.rda")

plot(mgamm_early_spei_noSp$gam,pages=1)
summary(mgamm_early_spei_noSp$lme) # details of underlying lme fit
summary(mgamm_early_spei_noSp$gam) # gam style summary of fitted model
anova(mgamm_early_spei_noSp$gam) 
gam.check(mgamm_early_spei_noSp$gam) # simple checking plots

# avec log
mgamm_early_log_spei_noSp <- gamm(cbind(JUV, AD) ~ s(log(meanSPEI_early+0.1)), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_log_spei_noSp, file="C:/git/STOC/stats/models/mgamm_early_log_spei_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_log_spei_noSp")

plot(mgamm_early_log_spei_noSp$gam,pages=1)
summary(mgamm_early_log_spei_noSp$lme) # details of underlying lme fit
summary(mgamm_early_log_spei_noSp$gam) # gam style summary of fitted model
anova(mgamm_early_log_spei_noSp$gam) 
gam.check(mgamm_early_log_spei_noSp$gam) # simple checking plots


#### BAL  

# plot BAL
sp <- ggplot(data=var_mod, aes(x=meanBAL_early, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
# sp <- sp + geom_abline(intercept = coeff[2], slope = coeff[1], color = "red", linewidth =1.5)+
  # ggtitle(eq)
plot(sp)

# Model
mgamm_early_BAL_noSp <- gamm(cbind(JUV, AD) ~ s(meanBAL_early), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_BAL_noSp, file="C:/git/STOC/stats/models/mgamm_early_BAL_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_BAL_noSp.rda")

plot(mgamm_early_BAL_noSp$gam,pages=1)
summary(mgamm_early_BAL_noSp$lme) # details of underlying lme fit
summary(mgamm_early_BAL_noSp$gam) # gam style summary of fitted model
anova(mgamm_early_BAL_noSp$gam) 
gam.check(mgamm_early_BAL_noSp$gam) # simple checking plots

# avec log
mgamm_early_log_BAL_noSp <- gamm(cbind(JUV, AD) ~ s(log(meanBAL_early+0.1)), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_log_BAL_noSp, file="C:/git/STOC/stats/models/mgamm_early_log_BAL_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_log_BAL_noSp")

plot(mgamm_early_log_BAL_noSp$gam,pages=1)
summary(mgamm_early_log_BAL_noSp$lme) # details of underlying lme fit
summary(mgamm_early_log_BAL_noSp$gam) # gam style summary of fitted model
anova(mgamm_early_log_BAL_noSp$gam) 
gam.check(mgamm_early_log_BAL_noSp$gam) # simple checking plots

####################
#### BAL + SPEI ####
####################

# Model
mgamm_early_BAL_SPEI_noSp <- gamm(cbind(JUV, AD) ~ s(meanBAL_early) + s(meanSPEI_early), random = list(YEAR = ~1, ID_PROG = ~1), family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_BAL_SPEI_noSp, file="C:/git/STOC/stats/models/mgamm_early_BAL_SPEI_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_BAL_SPEI_noSp.rda")

plot(mgamm_early_BAL_SPEI_noSp$gam,pages=1)
summary(mgamm_early_BAL_SPEI_noSp$lme) # details of underlying lme fit
summary(mgamm_early_BAL_SPEI_noSp$gam) # gam style summary of fitted model
anova(mgamm_early_BAL_SPEI_noSp$gam) 
gam.check(mgamm_early_BAL_SPEI_noSp$gam) # simple checking plots

# avec log
mgamm_early_log_BAL_SPEI_noSp <- gamm(cbind(JUV, AD) ~ s(meanBAL_early) + s(log(meanSPEI_early + 0.1)), random = list(YEAR = ~1, ID_PROG = ~1), family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_log_BAL_SPEI_noSp, file="C:/git/STOC/stats/models/mgamm_early_log_BAL_SPEI_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_log_BAL_SPEI_noSp")

plot(mgamm_early_log_BAL_SPEI_noSp$gam,pages=1)
summary(mgamm_early_log_BAL_SPEI_noSp$lme) # details of underlying lme fit
summary(mgamm_early_log_BAL_SPEI_noSp$gam) # gam style summary of fitted model
anova(mgamm_early_log_BAL_SPEI_noSp$gam) 
gam.check(mgamm_early_log_BAL_SPEI_noSp$gam) # simple checking plots

```

_late spring_  

```{r}
####  SPEI  ####
# plot SPEI
sp <- ggplot(data=var_mod, aes(x=meanSPEI_late, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

# Model

mgamm_late_spei <- gamm(cbind(JUV, AD) ~ s(meanSPEI_late), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod2000)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_spei, file="C:/git/STOC/stats/models/mgamm_late_spei.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_spei_noSp.rda")

plot(mgamm_late_spei$gam,pages=1)
summary(mgamm_late_spei$lme) # details of underlying lme fit
summary(mgamm_late_spei$gam) # gam style summary of fitted model
anova(mgamm_late_spei$gam) 
gam.check(mgamm_late_spei$gam) # simple checking plots


mgamm_late_spei_noSp <- gamm(cbind(JUV, AD) ~ s(meanSPEI_late), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_spei_noSp, file="C:/git/STOC/stats/models/mgamm_late_spei_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_spei_noSp.rda")

plot(mgamm_late_spei_noSp$gam,pages=1)
summary(mgamm_late_spei_noSp$lme) # details of underlying lme fit
summary(mgamm_late_spei_noSp$gam) # gam style summary of fitted model
anova(mgamm_late_spei_noSp$gam) 
gam.check(mgamm_late_spei_noSp$gam) # simple checking plots


####  BAL  ####
# plot BAL
sp <- ggplot(data=var_mod, aes(x=meanBAL_late, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
# sp <- sp + geom_abline(intercept = coeff[2], slope = coeff[1], color = "red", linewidth =1.5)+
  # ggtitle(eq)
plot(sp)

# Model
mgamm_late_BAL <- gamm(cbind(JUV, AD) ~ s(meanBAL_late), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod2000)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_BAL, file="C:/git/STOC/stats/models/mgamm_late_BAL.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_BAL_noSp.rda")

plot(mgamm_late_BAL$gam,pages=1)
summary(mgamm_late_BAL$lme) # details of underlying lme fit
summary(mgamm_late_BAL$gam) # gam style summary of fitted model
anova(mgamm_late_BAL$gam) 
gam.check(mgamm_late_BAL$gam) # simple checking plots


mgamm_late_BAL_noSp <- gamm(cbind(JUV, AD) ~ s(meanBAL_late), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_BAL_noSp, file="C:/git/STOC/stats/models/mgamm_late_BAL_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_BAL_noSp.rda")

plot(mgamm_late_BAL_noSp$gam,pages=1)
summary(mgamm_late_BAL_noSp$lme) # details of underlying lme fit
summary(mgamm_late_BAL_noSp$gam) # gam style summary of fitted model
anova(mgamm_late_BAL_noSp$gam) 
gam.check(mgamm_late_BAL_noSp$gam) # simple checking plots


####################
#### BAL + SPEI ####
####################
mgamm_late_BAL_SPEI_noSp <- gamm(cbind(JUV, AD) ~ s(log(meanSPEI_late + 0.1)) + s(meanBAL_late), random = list(YEAR = ~1, ID_PROG = ~1), family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_BAL_SPEI_noSp, file="C:/git/STOC/stats/models/mgamm_late_BAL_SPEI_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_BAL_SPEI_noSp.rda")

plot(mgamm_late_BAL_SPEI_noSp$gam,pages=1)
summary(mgamm_late_BAL_SPEI_noSp$lme) # details of underlying lme fit
summary(mgamm_late_BAL_SPEI_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_late_BAL_SPEI_noSp$gam) 
gam.check(mgamm_late_BAL_SPEI_noSp$gam) # simple checking plots

```


#### Temperature  

_Early spring_  
```{r}
prod <- fread("C:/git/STOC/Variables/data/model/Prod.csv")
var <- fread("C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv")

var_mod <- prod %>% left_join(var) %>% 
  filter(!is.na(meanSPEI_early)) # now we keep only our 33 species

# plot 
sp <- ggplot(data=var_mod, aes(x=meanT_early, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

mgamm_early_temp_noSp <- gamm(cbind(JUV, AD) ~ s(meanT_early), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_temp_noSp, file="C:/git/STOC/stats/models/mgamm_early_temp_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_temp_noSp.rda")

plot(mgamm_early_temp_noSp$gam,pages=1)
summary(mgamm_early_temp_noSp$lme) # details of underlying lme fit
summary(mgamm_early_temp_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_early_temp_noSp$gam) 
gam.check(mgamm_early_temp_noSp$gam) # simple checking plots

```


_Late spring_  
```{r}

# plot 
sp <- ggplot(data=var_mod, aes(x=meanT_late, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

mgamm_late_temp_noSp <- gamm(cbind(JUV, AD) ~ s(meanT_late), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_temp_noSp, file="C:/git/STOC/stats/models/mgamm_late_temp_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_temp_noSp.rda")

plot(mgamm_late_temp_noSp$gam,pages=1)
summary(mgamm_late_temp_noSp$lme) # details of underlying lme fit
summary(mgamm_late_temp_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_late_temp_noSp$gam) 
gam.check(mgamm_late_temp_noSp$gam) # simple checking plots

```


#### Temperature anomalies  

_Early spring_  
```{r}
prod <- fread("C:/git/STOC/Variables/data/model/Prod.csv")
var <- fread("C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv")

var_mod <- prod %>% left_join(var) %>% 
  filter(!is.na(meanSPEI_early)) # now we keep only our 33 species

# plot 
sp <- ggplot(data=var_mod, aes(x=mean_aT_early, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

mgamm_early_atemp <- gamm(cbind(JUV, AD) ~ s(mean_aT_early), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), 
                        family = binomial, data = var_mod2000)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_atemp, file="C:/git/STOC/stats/models/mgamm_early_atemp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_atemp.rda")

plot(mgamm_early_atemp$gam,pages=1)
summary(mgamm_early_atemp$lme) # details of underlying lme fit
summary(mgamm_early_atemp$gam) # gam style summary of fitted model
anova(mgamm_early_atemp$gam) 
gam.check(mgamm_early_atemp$gam) # simple checking plots


mgamm_early_atemp_noSp <- gamm(cbind(JUV, AD) ~ s(mean_aT_early), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_atemp_noSp, file="C:/git/STOC/stats/models/mgamm_early_atemp_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_atemp_noSp.rda")

plot(mgamm_early_atemp_noSp$gam,pages=1)
summary(mgamm_early_atemp_noSp$lme) # details of underlying lme fit
summary(mgamm_early_atemp_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_early_atemp_noSp$gam) 
gam.check(mgamm_early_atemp_noSp$gam) # simple checking plots

```


_Late spring_  
```{r}

# plot 
sp <- ggplot(data=var_mod, aes(x=mean_aT_late, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)


mgamm_late_atemp <- gamm(cbind(JUV, AD) ~ s(mean_aT_late), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), 
                        family = binomial, data = var_mod2000)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_atemp, file="C:/git/STOC/stats/models/mgamm_late_atemp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_atemp.rda")

plot(mgamm_late_atemp$gam,pages=1)
summary(mgamm_late_atemp$lme) # details of underlying lme fit
summary(mgamm_late_atemp$gam) # gam style summary of fitted model
anova(mgamm_late_atemp$gam) 
gam.check(mgamm_late_atemp$gam) # simple checking plots


mgamm_late_atemp_noSp <- gamm(cbind(JUV, AD) ~ s(mean_aT_late), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_atemp_noSp, file="C:/git/STOC/stats/models/mgamm_late_atemp_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_atemp_noSp.rda")

plot(mgamm_late_atemp_noSp$gam,pages=1)
summary(mgamm_late_atemp_noSp$lme) # details of underlying lme fit
summary(mgamm_late_atemp_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_late_atemp_noSp$gam) 
gam.check(mgamm_late_atemp_noSp$gam) # simple checking plots

```



#### NDVI  

_Early spring_  
```{r}
prod <- fread("C:/git/STOC/Variables/data/model/Prod.csv")
var <- fread("C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv")

var_mod <- prod %>% left_join(var) %>% 
  filter(!is.na(meanNDVI_early)) # now we keep only our 33 species

# plot 
sp <- ggplot(data=var_mod, aes(x=meanNDVI_early, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

mgamm_early_NDVI <- gamm(cbind(JUV, AD) ~ s(meanNDVI_early), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_NDVI, file="C:/git/STOC/stats/models/mgamm_early_NDVI.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_NDVI_noSp.rda")

plot(mgamm_early_NDVI$gam,pages=1)
summary(mgamm_early_NDVI$lme) # details of underlying lme fit
summary(mgamm_early_NDVI$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_early_NDVI$gam) 
gam.check(mgamm_early_NDVI$gam) # simple checking plots

```


_Late spring_  
```{r}

# plot 
sp <- ggplot(data=var_mod, aes(x=meanNDVI_late, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

mgamm_late_NDVI_noSp <- gamm(cbind(JUV, AD) ~ s(meanNDVI_late), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_NDVI_noSp, file="C:/git/STOC/stats/models/mgamm_late_NDVI_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_NDVI_noSp.rda")

plot(mgamm_late_NDVI_noSp$gam,pages=1)
summary(mgamm_late_NDVI_noSp$lme) # details of underlying lme fit
summary(mgamm_late_NDVI_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_late_NDVI_noSp$gam) 
gam.check(mgamm_late_NDVI_noSp$gam) # simple checking plots

```


### NDVI anomalies    

_Early spring_  
```{r}
prod <- fread("C:/git/STOC/Variables/data/model/Prod.csv")
var <- fread("C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv")

var_mod <- prod %>% left_join(var) %>% 
  filter(!is.na(meanNDVI_early)) # now we keep only our 33 species

# plot 
sp <- ggplot(data=var_mod, aes(x=mean_aNDVI_early, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

mgamm_early_aNDVI_noSp <- gamm(cbind(JUV, AD) ~ s(mean_aNDVI_early), random = list(YEAR = ~1, ID_PROG = ~1,  ESPECE = ~1),  family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_aNDVI_noSp, file="C:/git/STOC/stats/models/mgamm_early_aNDVI_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_aNDVI_noSp.rda")

plot(mgamm_early_aNDVI_noSp$gam,pages=1)
summary(mgamm_early_aNDVI_noSp$lme) # details of underlying lme fit
summary(mgamm_early_aNDVI_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_early_aNDVI_noSp$gam) 
gam.check(mgamm_early_aNDVI_noSp$gam) # simple checking plots

```

_Late spring_  
```{r}

# plot 
sp <- ggplot(data=var_mod, aes(x=mean_aNDVI_late, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)

mgamm_late_aNDVI_noSp <- gamm(cbind(JUV, AD) ~ s(mean_aNDVI_late), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_aNDVI_noSp, file="C:/git/STOC/stats/models/mgamm_late_aNDVI_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_aNDVI_noSp.rda")

plot(mgamm_late_aNDVI_noSp$gam,pages=1)
summary(mgamm_late_aNDVI_noSp$lme) # details of underlying lme fit
summary(mgamm_late_aNDVI_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_late_aNDVI_noSp$gam) 
gam.check(mgamm_late_aNDVI_noSp$gam) # simple checking plots

```



#### Temperature anomalies  

_Early spring_  
```{r}
prod <- fread("C:/git/STOC/Variables/data/model/Prod.csv")
var <- fread("C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv")

# plot 
sp <- ggplot(data=var_mod, aes(x=mean_aT_early, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)


var_mod <- prod %>% left_join(var) %>% 
  filter(!is.na(meanT_early)) # now we keep only our 33 species

mgamm_early_aT_noSp <- gamm(cbind(JUV, AD) ~ s(mean_aT_early), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_aT_noSp, file="C:/git/STOC/stats/models/mgamm_early_aT_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_aT_noSp.rda")

plot(mgamm_early_aT_noSp$gam,pages=1)
summary(mgamm_early_aT_noSp$lme) # details of underlying lme fit
summary(mgamm_early_aT_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_early_aT_noSp$gam) 
gam.check(mgamm_early_aT_noSp$gam) # simple checking plots

```


_Late spring_  
```{r}

# plot 
sp <- ggplot(data=var_mod, aes(x=mean_aT_late, y=Prod)) + geom_point(size = 0.7)
sp <- sp + geom_smooth(se=TRUE)
plot(sp)


mgamm_late_aT_noSp <- gamm(cbind(JUV, AD) ~ s(mean_aT_late), random = list(YEAR = ~1, ID_PROG = ~1), 
                        family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_aT_noSp, file="C:/git/STOC/stats/models/mgamm_late_aT_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_aT_noSp.rda")

plot(mgamm_late_aT_noSp$gam,pages=1)
summary(mgamm_late_aT_noSp$lme) # details of underlying lme fit
summary(mgamm_late_aT_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_late_aT_noSp$gam) 
gam.check(mgamm_late_aT_noSp$gam) # simple checking plots

```

#### temperature + anomalies  
_Early spring_
```{r}
prod <- fread("C:/git/STOC/Variables/data/model/Prod.csv")
var <- fread("C:/git/STOC/Variables/data/model/mean_var_sp_2periods.csv")

var_mod <- prod %>% left_join(var) %>% 
  filter(!is.na(meanSPEI_early)) # now we keep only our 33 species

mgamm_early_temp_a_noSp <- gamm(cbind(JUV, AD) ~ s(meanT_early) + s(mean_aT_early), random = list(YEAR = ~1, ID_PROG = ~1), family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_early_temp_a_noSp, file="C:/git/STOC/stats/models/mgamm_early_temp_a_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_early_temp_a_noSp.rda")

plot(mgamm_early_temp_a_noSp$gam,pages=1)
summary(mgamm_early_temp_a_noSp$lme) # details of underlying lme fit
summary(mgamm_early_temp_a_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_early_temp_a_noSp$gam) 
gam.check(mgamm_early_temp_a_noSp$gam) # simple checking plots

```

_late spring_  
```{r}

mgamm_late_temp_a_noSp <- gamm(cbind(JUV, AD) ~ s(meanT_late) + s(mean_aT_late), random = list(YEAR = ~1, ID_PROG = ~1), family = binomial, data = var_mod)  
# This model takes a long time to run, therefore it is saved for later use:
save(mgamm_late_temp_a_noSp, file="C:/git/STOC/stats/models/mgamm_late_temp_a_noSp.rda", compress='xz')
# load the model:
# load("C:/git/STOC/stats/models/mgamm_late_temp_a_noSp.rda")

plot(mgamm_late_temp_a_noSp$gam,pages=1)
summary(mgamm_late_temp_a_noSp$lme) # details of underlying lme fit
summary(mgamm_late_temp_a_noSp$gam) # gam style summary of fitted model
# It's the same whether the variable is scaled or not
anova(mgamm_late_temp_a_noSp$gam) 
gam.check(mgamm_late_temp_a_noSp$gam) # simple checking plots

```



## Use a single date for all years and species  
It is good to have a precise phenology per year and species, but it still limits the number of species we include in the analysis. So I will also use a mean or median hatching date to determine the early and late periods.  

```{r}
# Import pheno dates  
pheno_dates <- fread("C:/git/STOC/Variables/data/pheno_hatching.csv")
summary(pheno_dates)

hist(pheno_dates$Hatching, breaks = 15)

mean_hatching <- mean(pheno_dates$Hatching) 
# 132.5  ;  i.e. 12/05
median_hatching <- median(pheno_dates$Hatching) 
# 134  ;  i.e. 14/05

```

I will use the median hatching date 14/05 (Julian day 134) as a limit between early and late breeding period.  

I computed early and late spring SPEI in the file SPEI.Rmd  

```{r SPEI 2 periods}
# import early SPEI
early_spei <- fread("C:/git/STOC/Variables/data/early_spei_month2.csv")
summary(early_spei)

# import late SPEI
late_spei <- fread("C:/git/STOC/Variables/data/late_spei_month2.csv")
summary(late_spei)

# plot
hist(early_spei$SPEI) # normal
hist(late_spei$SPEI) # normal  

```

Early and late SPEI have a similar distribution, normal and centerd on 0.  

### Build dataset with all variables  

```{r early late variables}

# Import climate data  
climate <- fread("C:/git/STOC/Variables/data/model/var_climat_daily_anomalies.csv")
# Import NDVI
ndvi <- fread("C:/git/STOC/Variables/data/model/var_NDVI_daily_anomalies.csv")

# keep only useful climate data  
climate_pheno <- climate %>% 
  rename(YEAR = an) %>% 
  filter(d > 72 & d <197) %>% # keep spring data
  dplyr::select(ID_PROG, Date, d, YEAR, T_Q, aTemp, BAL) 

# Create variable "early" and "late"  
climate_pheno <- climate_pheno %>% 
  mutate(period = ifelse(d < 134, "early", "late")) %>%  # early until median hatching date, day = 134
  group_by(ID_PROG, YEAR, period) %>% 
  summarize(meanBAL = mean(BAL), meanT = mean(T_Q), mean_aT = mean(aTemp))
  
# wider  
climate_pheno_wide <- climate_pheno %>% 
  filter(!is.na(period)) %>% 
  pivot_wider(names_from = period, values_from = c(meanBAL, meanT, mean_aT))

# Merge with SPEI
early_spei <- early_spei %>% mutate(early_SPEI = SPEI) %>% select(Year, ID_PROG, early_SPEI) %>% 
  mutate(ID_PROG = as.factor(ID_PROG))
late_spei <- late_spei %>% mutate(late_SPEI = SPEI) %>% select(Year, ID_PROG, late_SPEI) %>% 
  mutate(ID_PROG = as.factor(ID_PROG))

climate_pheno_wide <- climate_pheno_wide %>% 
  mutate(ID_PROG = as.factor(ID_PROG)) %>%
  left_join(early_spei, by = c("ID_PROG", "YEAR" = "Year")) %>% 
  left_join(late_spei, by = c("ID_PROG", "YEAR" = "Year"))

# write.csv(climate_pheno_wide, file = "C:/git/STOC/Variables/data/model/mean_climate_2periods.csv", row.names = FALSE)

# Merge NDVI with dates  
ndvi_pheno <- ndvi %>% 
  filter(d > 72 & d <197) %>% # keep spring data (normalement on a déjà que le printemps mais bon)
  dplyr::select(ID_PROG, Date, d, YEAR, NDVI, daNDVI) %>% 
# Create variables "early" and "late"  
  mutate(period = ifelse(d < 134, "early", "late")) %>% 
  group_by(ID_PROG, YEAR, period) %>% 
  summarize(meanNDVI = mean(NDVI, na.rm = TRUE), mean_aNDVI = mean(daNDVI, na.rm = TRUE)) %>%  
  filter(!is.na(period)) %>% 
  pivot_wider(names_from = period, values_from = c(meanNDVI, mean_aNDVI))

# write.csv(ndvi_pheno, file = "C:/git/STOC/Variables/data/model/mean_ndvi_2periods.csv", row.names = FALSE)

# merge df  
mean_2periods <- merge(climate_pheno_wide, ndvi_pheno, all = TRUE)
# We only have data from 2000 for NDVI

# write.csv(mean_2periods, file = "C:/git/STOC/Variables/data/model/mean_var_2periods.csv", row.names = FALSE)

```

