---
title: "Models"
author: "Nathalie Adenot"
date: "2023-03-01"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r packages, echo=FALSE}
require(tidyverse)
require(glmmTMB)
require(data.table)
require(DHARMa)
library(mgcv)

#libraries for spatial series analyses
library(spdep) #to extracted the neighbors is a spatial data
library(ade4)  #used for plotting spatial data
library(spatialreg) #used for spatial data modelling
library(gwrr)  #to run geographically weighted regression

```

 
 
# GAMM  

We will use the function *gamm* from the *mgcv* package.  

I start with univariate models without interactions to explore the linearity of relationships between environmental variables and productivity.  


_Load data_  
For the final model, we decided to keep mean spring variables for Temperature, water balance and NDVI.  
And early / late variables for anomalies: SPEI, temperature anomalies, NDVI anomalies  

```{r data}
data_mod <- fread("C:/git/STOC/Variables/data/model/df_model_final_ECE.csv")

data_mod <- data_mod %>% 
  mutate(ID_PROG = as.factor(ID_PROG)) %>% 
  # mutate(YEAR = as.factor(YEAR)) %>% 
  mutate(ESPECE = as.factor(ESPECE))
```

_subset dataset_  
As is requires a lot of RAM to run these GAMM with 3 random effects, I need to subset the dataset.  
I will keep only the 50 most captured species, and data from 2000.  
It shouldn't bias the results and we still have quite a lot of data.  

```{r subset data}

# list of species  
pheno <- fread("C:/git/STOC/Variables/data/pheno.csv")
species <- unique(pheno$ESPECE)

# subset data_mod  
data_mod_50sp <- data_mod %>% 
  filter(ESPECE %in% species) %>% # keep 50 species
  filter(YEAR > 2000) # keep data from 2000


```

## SPEI  

### early SPEI
```{r early SPEI}

gamm_early_spei <- gamm(cbind(JUV, AD) ~ s(early_SPEI), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_early_spei, file="C:/git/STOC/stats/models/gamm_early_spei.rda", compress='xz')
load("C:/Users/Nathalie Adenot/STOC/cluster/output/gamm_early_spei.rda")

plot(gamm_early_spei$gam,pages=1)
summary(gamm_early_spei$lme) # details of underlying lme fit
summary(gamm_early_spei$gam) # gam style summary of fitted model
anova(gamm_early_spei$gam) 
gam.check(gamm_early_spei$gam) # simple checking plots

```

### late SPEI  

```{r late SPEI}

gamm_late_spei <- gamm(cbind(JUV, AD) ~ s(late_SPEI), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_late_spei, file="C:/git/STOC/stats/models/gamm_late_spei.rda", compress='xz')

plot(gamm_late_spei$gam,pages=1)
summary(gamm_late_spei$lme) # details of underlying lme fit
summary(gamm_late_spei$gam) # gam style summary of fitted model
anova(gamm_late_spei$gam) 
gam.check(gamm_late_spei$gam) # simple checking plots

```

## Water balance (BAL)  

```{r BAL}
gamm_BAL <- gamm(cbind(JUV, AD) ~ s(mean_BAL), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_BAL, file="C:/git/STOC/stats/models/gamm_BAL.rda", compress='xz')

plot(gamm_BAL$gam,pages=1)
summary(gamm_BAL$lme) # details of underlying lme fit
summary(gamm_BAL$gam) # gam style summary of fitted model
anova(gamm_BAL$gam) 
gam.check(gamm_BAL$gam) # simple checking plots

```


## NDVI  

As there are NAs in the data, we are able to run the GAMM on the complete dataset.  

### Mean NDVI  

```{r mean NDVI}

gamm_NDVI <- gamm(cbind(JUV, AD) ~ s(mean_NDVI), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod) 

save(gamm_NDVI, file="C:/git/STOC/stats/models/gamm_NDVI.rda", compress='xz')

plot(gamm_NDVI$gam,pages=1)
summary(gamm_NDVI$lme) # details of underlying lme fit
summary(gamm_NDVI$gam) # gam style summary of fitted model
anova(gamm_NDVI$gam) 
gam.check(gamm_NDVI$gam) # simple checking plots

```

### NDVI anomalies  

_Early breeding season_  

```{r early NDVI anomalies}

gamm_early_aNDVI <- gamm(cbind(JUV, AD) ~ s(mean_aNDVI_early), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod) 

save(gamm_early_aNDVI, file="C:/git/STOC/stats/models/gamm_early_aNDVI.rda", compress='xz')

plot(gamm_early_aNDVI$gam,pages=1)
summary(gamm_early_aNDVI$lme) # details of underlying lme fit
summary(gamm_early_aNDVI$gam) # gam style summary of fitted model
anova(gamm_early_aNDVI$gam) 
gam.check(gamm_early_aNDVI$gam) # simple checking plots

```


_Late breeding season_  

```{r late NDVI anomalies}

gamm_late_aNDVI <- gamm(cbind(JUV, AD) ~ s(mean_aNDVI_late), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod) 

save(gamm_late_aNDVI, file="C:/git/STOC/stats/models/gamm_late_aNDVI.rda", compress='xz')

plot(gamm_late_aNDVI$gam,pages=1)
summary(gamm_late_aNDVI$lme) # details of underlying lme fit
summary(gamm_late_aNDVI$gam) # gam style summary of fitted model
anova(gamm_late_aNDVI$gam) 
gam.check(gamm_late_aNDVI$gam) # simple checking plots

```



## Temperature  

### Mean temperature  

```{r mean temperature}

gamm_T <- gamm(cbind(JUV, AD) ~ s(mean_T), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_T, file="C:/git/STOC/stats/models/gamm_T.rda", compress='xz')

plot(gamm_T$gam,pages=1)
summary(gamm_T$lme) # details of underlying lme fit
summary(gamm_T$gam) # gam style summary of fitted model
anova(gamm_T$gam) 
gam.check(gamm_T$gam) # simple checking plots

```

### Temperature anomalies  

_Early breeding season_  

```{r early Temperature anomalies}

gamm_early_aT <- gamm(cbind(JUV, AD) ~ s(mean_aT_early), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_early_aT, file="C:/git/STOC/stats/models/gamm_early_aT.rda", compress='xz')

plot(gamm_early_aT$gam,pages=1)
summary(gamm_early_aT$lme) # details of underlying lme fit
summary(gamm_early_aT$gam) # gam style summary of fitted model
anova(gamm_early_aT$gam) 
gam.check(gamm_early_aT$gam) # simple checking plots

```

_Late breeding season_  

```{r late Temperature anomalies}

gamm_late_aT <- gamm(cbind(JUV, AD) ~ s(mean_aT_late), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_late_aT, file="C:/git/STOC/stats/models/gamm_late_aT.rda", compress='xz')

plot(gamm_late_aT$gam,pages=1)
summary(gamm_late_aT$lme) # details of underlying lme fit
summary(gamm_late_aT$gam) # gam style summary of fitted model
anova(gamm_late_aT$gam) 
gam.check(gamm_late_aT$gam) # simple checking plots

```


## ECE  

For SPEI and Temperature, I defined several variables to characterise ECE (Extreme Climatic Events):  
- *ECE_spei* & *ECE_temp*: binary variables (true/false), stating whether a year is extreme or not  
- *nb_ECE_SPEI* & *nb_ECE_aTemp*: continuous variable, giving the number of ECE days per year  
- *ECE3temp* & *ECE3SPEI*: continuous variables, giving the number of times per year there were more than 3 consecutive ECE days  


_ECE SPEI_    

```{r ECE SPEI}

gamm_ECE_spei <- gamm(cbind(JUV, AD) ~ s(ECE_spei), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_ECE_spei, file="C:/git/STOC/stats/models/gamm_ECE_spei.rda", compress='xz')

plot(gamm_ECE_spei$gam,pages=1)
summary(gamm_ECE_spei$lme) # details of underlying lme fit
summary(gamm_ECE_spei$gam) # gam style summary of fitted model
anova(gamm_ECE_spei$gam) 
gam.check(gamm_ECE_spei$gam) # simple checking plots

```

_ECE temperature_    

```{r ECE temperature}

gamm_ECE_temp <- gamm(cbind(JUV, AD) ~ s(ECE_temp), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_ECE_temp, file="C:/git/STOC/stats/models/gamm_ECE_temp.rda", compress='xz')

plot(gamm_ECE_temp$gam,pages=1)
summary(gamm_ECE_temp$lme) # details of underlying lme fit
summary(gamm_ECE_temp$gam) # gam style summary of fitted model
anova(gamm_ECE_temp$gam) 
gam.check(gamm_ECE_temp$gam) # simple checking plots

```


_nb ECE SPEI_    

```{r nb ECE SPEI}

gamm_nb_ECE_spei <- gamm(cbind(JUV, AD) ~ s(nb_ECE_SPEI), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_nb_ECE_spei, file="C:/git/STOC/stats/models/gamm_nb_ECE_SPEI.rda", compress='xz')

plot(gamm_nb_ECE_spei$gam,pages=1)
summary(gamm_nb_ECE_spei$lme) # details of underlying lme fit
summary(gamm_nb_ECE_spei$gam) # gam style summary of fitted model
anova(gamm_nb_ECE_spei$gam) 
gam.check(gamm_nb_ECE_spei$gam) # simple checking plots

```

_nb ECE temperature_    

```{r nb ECE aTemp}

gamm_nb_ECE_aTemp <- gamm(cbind(JUV, AD) ~ s(nb_ECE_aTemp), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_nb_ECE_aTemp, file="C:/git/STOC/stats/models/gamm_nb_ECE_aTemp.rda", compress='xz')

plot(gamm_nb_ECE_aTemp$gam,pages=1)
summary(gamm_nb_ECE_aTemp$lme) # details of underlying lme fit
summary(gamm_nb_ECE_aTemp$gam) # gam style summary of fitted model
anova(gamm_nb_ECE_aTemp$gam) 
gam.check(gamm_nb_ECE_aTemp$gam) # simple checking plots

```


_nb 3 ECE SPEI_  

```{r nb 3 ECE cons SPEI}

gamm_nb_ECE3SPEI <- gamm(cbind(JUV, AD) ~ s(ECE3SPEI), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_nb_ECE3SPEI, file="C:/git/STOC/stats/models/gamm_nb_ECE3SPEI.rda", compress='xz')

plot(gamm_nb_ECE3SPEI$gam,pages=1)
summary(gamm_nb_ECE3SPEI$lme) # details of underlying lme fit
summary(gamm_nb_ECE3SPEI$gam) # gam style summary of fitted model
anova(gamm_nb_ECE3SPEI$gam) 
gam.check(gamm_nb_ECE3SPEI$gam) # simple checking plots

```

_nb 3 ECE temp_  

```{r nb 3 ECE cons temperature}

gamm_nb_ECE3temp <- gamm(cbind(JUV, AD) ~ s(ECE3temp), random = list(YEAR = ~1, ID_PROG = ~1, ESPECE = ~1), family = binomial, data = data_mod_50sp) 

save(gamm_nb_ECE3temp, file="C:/git/STOC/stats/models/gamm_nb_ECE3temp.rda", compress='xz')

plot(gamm_nb_ECE3temp$gam,pages=1)
summary(gamm_nb_ECE3temp$lme) # details of underlying lme fit
summary(gamm_nb_ECE3temp$gam) # gam style summary of fitted model
anova(gamm_nb_ECE3temp$gam) 
gam.check(gamm_nb_ECE3temp$gam) # simple checking plots

```






## ggpredict  

Now that I have explored the relationships of the environmental variables with productivity, I will check how good would be the use of simplified functions (linear or quadratic) to approximate these relationships.  

### SPEI  
_Early SPEI_  

```{r ggpredict SPEI}
require(ggeffects)
require(glmmTMB)
require(lme4)

##  GAMM
load("C:/git/STOC/stats/models/mgamm_early_spei.rda") # mean daily SPEI
load("C:/git/STOC/stats/models/gamm_early_spei.rda") # monthly SPEI computed over 2 months

# plot
# mean daily SPEI
ggpredict(mgamm_early_spei, 
          terms = "meanSPEI_early") |> 
  plot()
# monthly SPEI computed over 2 months
ggpredict(gamm_early_spei, 
          terms = "early_SPEI") |> 
  plot()

# get values
df_early_spei <- ggpredict(mgamm_early_spei, 
          terms = "meanSPEI_early") |> as.data.frame()  

## Quadratic  
# data 
# mean daily SPEI
var_model <- fread("C:/git/STOC/Variables/data/model/data_model_periods.csv")
# monthly SPEI  
var_model <- fread("C:/git/STOC/Variables/data/model/df_model_final.csv")

mod_early_spei2 <- glmmTMB(cbind(JUV, AD) ~ meanSPEI_early + I(meanSPEI_early^2) +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)
# plot
ggpredict(mod_early_spei2, 
          terms = c("meanSPEI_early")) |> 
  plot()
# get values
df_early_spei_2 <- ggpredict(mod_early_spei2, 
          terms = "meanSPEI_early [all]") |> as.data.frame()  


## Linear  
mod_early_spei_lin <- glmmTMB(cbind(JUV, AD) ~ early_SPEI +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)
summary(mod_early_spei_lin)
# plot
ggpredict(mod_early_spei_lin, 
          terms = c("early_SPEI [all]")) |> 
  plot()
# get values
df_early_spei_lin <- ggpredict(mod_early_spei_lin, 
          terms = "early_SPEI [all]") |> as.data.frame()  

```

The quadratic estimate is very far from the smooth term...


_Late SPEI_  

```{r}
require(ggeffects)
require(glmmTMB)
require(lme4)

##  GAMM
load("C:/git/STOC/stats/models/mgamm_late_spei.rda") # mean daily SPEI
load("C:/git/STOC/stats/models/gamm_late_spei.rda") # monthly SPEI computed over 2 months

# plot
# mean daily SPEI
ggpredict(mgamm_late_spei, 
          terms = "meanSPEI_late") |> 
  plot()
# monthly SPEI computed over 2 months
ggpredict(gamm_late_spei, 
          terms = "late_SPEI") |> 
  plot()

# get values
df_late_spei <- ggpredict(mgamm_late_spei, 
          terms = "meanSPEI_late") |> as.data.frame()  

## Quadratic  
# data 
# mean daily SPEI
var_model <- fread("C:/git/STOC/Variables/data/model/data_model_periods.csv")
# monthly SPEI  
var_model <- fread("C:/git/STOC/Variables/data/model/df_model_final.csv")

mod_late_spei2 <- glmmTMB(cbind(JUV, AD) ~ meanSPEI_late + I(meanSPEI_late^2) +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)
# plot
ggpredict(mod_late_spei2, 
          terms = c("meanSPEI_late [all]")) |> 
  plot()
# get values
df_late_spei_2 <- ggpredict(mod_late_spei2, 
          terms = "meanSPEI_late [all]") |> as.data.frame()  


## Linear  
mod_late_spei_lin <- glmmTMB(cbind(JUV, AD) ~ late_SPEI +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)
summary(mod_late_spei_lin)
# plot
ggpredict(mod_late_spei_lin, 
          terms = c("late_SPEI [all]")) |> 
  plot()
# get values
df_late_spei_lin <- ggpredict(mod_late_spei_lin, 
          terms = "late_SPEI [all]") |> as.data.frame()  

```

Same, the quadratic term predicts opposite trends compared to the GAMM...  

### Water balance  
```{r ggpredict BAL}  
require(ggeffects)
require(glmmTMB)
require(lme4)

##  GAMM
load("C:/git/STOC/stats/models/gamm_BAL.rda")

# plot
ggpredict(gamm_BAL, 
          terms = "mean_BAL") |> 
  plot()
# get values
df_BAL_spei <- ggpredict(mgamm_BAL_spei, 
          terms = "mean_BAL") |> as.data.frame()  

## Quadratic  
# data 
var_model <- fread("C:/git/STOC/Variables/data/model/data_mod_50sp.csv")

# Add quadratic terms
# SPEI early + late,  temperature anomalies late
var_model <- var_model %>% 
  mutate(SPEI_early2 = meanSPEI_early * meanSPEI_early) %>% 
  mutate(SPEI_late2 = meanSPEI_late * meanSPEI_late) %>% 
  mutate(aT_late2 = mean_aT_late * mean_aT_late)

mod_early_spei2 <- glmmTMB(cbind(JUV, AD) ~ meanSPEI_early + I(meanSPEI_early^2) +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)


# plot
ggpredict(mod_early_spei2, 
          terms = c("meanSPEI_early [all]")) |> 
  plot()

# get values
df_early_spei_2 <- ggpredict(mod_early_spei2, 
          terms = "meanSPEI_early [all]") |> as.data.frame()  


```


### Temperature

```{r ggpredict temp continuous}
##  GAMM
load("C:/git/STOC/stats/models/gamm_T.rda")

# plot
ggpredict(gamm_T, 
          terms = "mean_T") |> 
  plot()
# get values
df_T <- ggpredict(gamm_T, 
          terms = "mean_T") |> as.data.frame()  


## Quadratic  
# data 
var_model <- fread("C:/git/STOC/Variables/data/model/data_mod_50sp.csv")

mod_T2 <- glmmTMB(cbind(JUV, AD) ~ mean_T + I(mean_T^2) +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)

# plot
ggpredict(mod_T2, 
          terms = c("mean_T [all]")) |> 
  plot()

# get values
df_T2 <- ggpredict(mod_T2, 
          terms = "mean_T [all]") |> as.data.frame()  

```

### Temperature anomalies  

_Early temperature anomalies_  

```{r}
require(ggeffects)
require(glmmTMB)
require(lme4)

##  GAMM
load("C:/git/STOC/stats/models/gamm_early_aT.rda")

# plot
ggpredict(gamm_aT_early, 
          terms = "mean_aT_early") |> 
  plot()
# get values
df_early_aT <- ggpredict(gamm_early_aT, 
          terms = "mean_aT_early") |> as.data.frame()  


## Linear  
# data 
var_model <- fread("C:/git/STOC/Variables/data/model/data_mod_50sp.csv")

mod_early_aT_lin <- glmmTMB(cbind(JUV, AD) ~ mean_aT_early +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)

# plot
ggpredict(mod_early_aT_lin, 
          terms = c("mean_aT_early [all]")) |> 
  plot()

# get values
df_early_aT_lin <- ggpredict(mod_early_aT_lin, 
          terms = "mean_aT_early [all]") |> as.data.frame()  

```

_Late temperature anomalies_  

```{r}
require(ggeffects)
require(glmmTMB)
require(lme4)

##  GAMM
load("C:/git/STOC/stats/models/gamm_late_aT.rda")

# plot
ggpredict(gamm_aT_late, 
          terms = "mean_aT_late") |> 
  plot()
# get values
df_late_aT <- ggpredict(gamm_late_aT, 
          terms = "mean_aT_late") |> as.data.frame()  


## Linear  
# data 
var_model <- fread("C:/git/STOC/Variables/data/model/data_mod_50sp.csv")

mod_late_aT_lin <- glmmTMB(cbind(JUV, AD) ~ mean_aT_late +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)

# plot
ggpredict(mod_late_aT_lin, 
          terms = c("mean_aT_late [all]")) |> 
  plot()

# get values
df_late_aT_lin <- ggpredict(mod_late_aT_lin, 
          terms = "mean_aT_late [all]") |> as.data.frame()  


## Quadratic 
mod_late_aT2 <- glmmTMB(cbind(JUV, AD) ~ mean_aT_late + I(mean_aT_late^2) +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)

# plot
ggpredict(mod_late_aT2, 
          terms = c("mean_aT_late [all]")) |> 
  plot()

# get values
df_late_aT2 <- ggpredict(mod_late_aT2, 
          terms = "mean_aT_late [all]") |> as.data.frame()  


```



### NDVI

```{r ggpredict NDVI}
##  GAMM
load("C:/git/STOC/stats/models/gamm_NDVI.rda")

# plot
ggpredict(gamm_NDVI, 
          terms = "meanNDVI") |> 
  plot()
# get values
df_NDVI <- ggpredict(gamm_NDVI, 
          terms = "meanNDVI") |> as.data.frame()  


## Quadratic  
# data 
var_model <- fread("C:/git/STOC/Variables/data/model/data_mod_50sp.csv")

mod_NDVI2 <- glmmTMB(cbind(JUV, AD) ~ meanNDVI + I(meanNDVI^2) +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)

# plot
ggpredict(mod_NDVI2, 
          terms = c("meanNDVI [all]")) |> 
  plot()

# get values
df_NDVI2 <- ggpredict(mod_NDVI2, 
          terms = "meanNDVI [all]") |> as.data.frame()  

## Linear  
mod_NDVI_lin <- glmmTMB(cbind(JUV, AD) ~ meanNDVI +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)
# plot
ggpredict(mod_NDVI_lin, 
          terms = c("meanNDVI [all]")) |> 
  plot()
# get values
df_NDVI_lin <- ggpredict(mod_NDVI_lin, 
          terms = "meanNDVI [all]") |> as.data.frame()  

```

### NDVI anomalies  

_Early NDVI anomalies_  

```{r}
require(ggeffects)
require(glmmTMB)
require(lme4)

##  GAMM
load("C:/git/STOC/stats/models/gamm_early_aNDVI.rda")

# plot
ggpredict(gamm_aNDVI_early, 
          terms = "mean_aNDVI_early") |> 
  plot()
# get values
df_early_aNDVI <- ggpredict(gamm_early_aNDVI, 
          terms = "mean_aNDVI_early") |> as.data.frame()  


## Linear  
# data 
var_model <- fread("C:/git/STOC/Variables/data/model/data_mod_50sp.csv")

mod_early_aNDVI_lin <- glmmTMB(cbind(JUV, AD) ~ mean_aNDVI_early +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)

# plot
ggpredict(mod_early_aNDVI_lin, 
          terms = c("mean_aNDVI_early [all]")) |> 
  plot()

# get values
df_early_aNDVI_lin <- ggpredict(mod_early_aNDVI_lin, 
          terms = "mean_aNDVI_early [all]") |> as.data.frame()  

## Quadratic 
mod_early_aNDVI2 <- glmmTMB(cbind(JUV, AD) ~ mean_aNDVI_early + I(mean_aNDVI_early^2) +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)
# plot
ggpredict(mod_early_aNDVI2, 
          terms = c("mean_aNDVI_early [all]")) |> 
  plot()
# get values
df_early_aNDVI2 <- ggpredict(mod_early_aNDVI2, 
          terms = "mean_aNDVI_early [all]") |> as.data.frame()  


```

_Late temperature anomalies_  

```{r}
require(ggeffects)
require(glmmTMB)
require(lme4)

##  GAMM
load("C:/git/STOC/stats/models/gamm_late_aNDVI.rda")

# plot
ggpredict(gamm_aNDVI_late, 
          terms = "mean_aNDVI_late") |> 
  plot()
# get values
df_late_aNDVI <- ggpredict(gamm_late_aNDVI, 
          terms = "mean_aNDVI_late") |> as.data.frame()  


## Linear  
# data 
var_model <- fread("C:/git/STOC/Variables/data/model/data_mod_50sp.csv")

mod_late_aNDVI_lin <- glmmTMB(cbind(JUV, AD) ~ mean_aNDVI_late +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)

# plot
ggpredict(mod_late_aNDVI_lin, 
          terms = c("mean_aNDVI_late [all]")) |> 
  plot()

# get values
df_late_aNDVI_lin <- ggpredict(mod_late_aNDVI_lin, 
          terms = "mean_aNDVI_late [all]") |> as.data.frame()  


## Quadratic 
mod_late_aNDVI2 <- glmmTMB(cbind(JUV, AD) ~ mean_aNDVI_late + I(mean_aNDVI_late^2) +
                (1|ID_PROG) + (1|YEAR) + (1|ESPECE),
              family=binomial, data=var_model)
# plot
ggpredict(mod_late_aNDVI2, 
          terms = c("mean_aNDVI_late [all]")) |> 
  plot()
# get values
df_late_aNDVI2 <- ggpredict(mod_late_aNDVI2, 
          terms = "mean_aNDVI_late [all]") |> as.data.frame()  


```

